---
title: "RQ1 and RQ2"
output: html_notebook
author: "Santiago del Rey"
---

## Load libraries
```{r include=FALSE}
library(dplyr)
library(arrow)
library(ARTool)
```

## Load data
We load the data from the parquet file. We filter out the `inception_v3` architecture and the `Local Normal User` training environment, as they are not relevant for our analysis. We also rename the `energy (MJ)` column to `energy` for convenience.
```{r}
df <- read_parquet("../data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
  filter(architecture != "inception_v3" & `training environment` != "Local Normal User") %>%
  mutate(
    energy=`energy (MJ)`,
    training_environment= factor(`training environment`),
    base_model=factor(architecture),
    subject=factor(case_when(
            architecture == "mobilenet_v2" ~ "model1",
            architecture == "nasnet_mobile" ~ "model2",
            architecture == "resnet50" ~ "model3",
            architecture == "xception" ~ "model4",
            architecture == "vgg16" ~ "model5",
          ))
    )
```

## Create the model
We create a linear model with `energy` as the response variable and `training_environment` and `base_model` as predictors. We also include an error term for the `subject` variable to account for the fact that it is a random effect.
```{r}
m <- art(energy ~ training_environment*base_model+Error(subject), data=df)
summary(m)
```

## Test normality of the residuals
We use the Shapiro-Wilk test to check the normality of the residuals.
```{r}
resutls <- shapiro.test(residuals(m))

if (resutls$p.value < 0.05) {
  cat("The residuals are not normally distributed (p-value:", resutls$p.value, ")\n")
} else {
  cat("The residuals are normally distributed (p-value:", resutls$p.value, ")\n")
}
```

```{r}
anova(m)
```

```{r}
art.con(m, "training_environment")
art.con(m, "training_environment:base_model") %>%
  summary() %>%
  mutate(sig = symnum(p.value, corr = FALSE, na = FALSE,
                      cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                      symbols = c("***", "**", "*", ".", " ")
  ))
```
