---
title: "RQ1: What is the impact of model architecture and training environment on image classification training concerning energy consumption and GPU usage?"
output: html_notebook
author: "Santiago del Rey"
---

```{r include=FALSE}
# Load libraries
library(arrow)
library(dplyr)
library(lme4)
library(lmerTest) # p-values for lmer
library(DHARMa)
library(emmeans) # post hoc tests
library(gridExtra)
library(ggplot2)
library(ggpubr) # for publication-ready plots
library(car)
library(MASS)
```

## Load and preprocess data

We load the data from the parquet file, filter out the `inception_v3` architecture and the `Desktop Normal User` training environment, and normalize the energy consumption by the total number of images seen during training (converting MJ to J). We also prepare the `gpu_usage` variable.

```{r}
df <- read_parquet("../data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
  filter(architecture != "inception_v3" & `training environment` != "Desktop Normal User") %>%
  mutate(
    energy = (`energy (MJ)` * 1000000) / `total seen images`, # Convert MJ to J and normalize by total seen images
    training_environment = factor(`training environment`, levels = c("Desktop ML Engineer", "Server")),
    architecture = case_match(architecture,
      "mobilenet_v2" ~ "Mobilenet V2",
      "nasnet_mobile" ~ "NASNet Mobile",
      "resnet50" ~ "ResNet50",
      "xception" ~ "Xception",
      "vgg16" ~ "VGG16"
    ),
    architecture = factor(architecture, levels = c("Mobilenet V2", "NASNet Mobile", "ResNet50", "Xception", "VGG16")),
    gpu_usage = `gpu usage (%)`
  ) %>%
  dplyr::select(energy, gpu_usage, architecture, training_environment)
```

### Outlier filtering functions

```{r}
# Define a function to filter outliers of a given variable on an input data frame
filter_outliers <- function(df, variable) {
  df %>%
    group_by(architecture, training_environment) %>%
    mutate(
      Q1 = quantile(!!sym(variable), 0.25),
      Q3 = quantile(!!sym(variable), 0.75),
      IQR = Q3 - Q1,
      is_outlier = !!sym(variable) < (Q1 - 1.5 * IQR) | !!sym(variable) > (Q3 + 1.5 * IQR)
    ) %>%
    ungroup() %>%
    filter(!is_outlier) %>%
    dplyr::select(-Q1, -Q3, -IQR, -is_outlier)
}
```

---

# Energy Consumption Analysis

## Data summary

```{r}
filtered_energy_data <- df %>%
  filter_outliers("energy")

filtered_energy_data %>%
  group_by(architecture, training_environment) %>%
  summarise(
    mean_energy = mean(energy, na.rm = TRUE),
    sd_energy = sd(energy, na.rm = TRUE),
    min_energy = min(energy, na.rm = TRUE),
    max_energy = max(energy, na.rm = TRUE),
    n = n()
  )
```

```{r}
hist(filtered_energy_data$energy, breaks = 30, main = "Distribution of Energy")
```

## Linear Model

Fit a linear model to the energy consumption data, including an interaction term between architecture and training environment.
```{r}
model_energy <- lm(energy ~ architecture * training_environment, data = filtered_energy_data)
summary(model_energy)
```

## Model diagnostics

```{r}
sim_res_energy <- simulateResiduals(fittedModel = model_energy)
res_df <- data.frame(
  scaled_residuals = sim_res_energy$scaledResiduals,
  predicted = sim_res_energy$fittedPredictedResponse,
  observed = sim_res_energy$observedResponse
)

testResiduals(sim_res_energy)
```

### Remove the outliers detected in the residuals and refit the model
The results of the outlier test indicate that there are some outliers in the residuals. We will remove these outliers based on the scaled residuals, which are the standardized residuals from the model.
We will filter out any observations with scaled residuals outside the 2.5%–97.5% quantiles.
```{r}
# Remove outliers based on the simulated residuals
scaled_res <- sim_res_energy$scaledResiduals

# Identify outliers: residuals outside the 2.5%–97.5% quantiles
lower <- quantile(scaled_res, 0.025)
upper <- quantile(scaled_res, 0.975)

# Mark which residuals are outside the acceptable range
outlier_index <- which(scaled_res < lower | scaled_res > upper)

filtered_energy_data <- filtered_energy_data[-outlier_index, ]

# Show the outliers
filtered_energy_data[outlier_index, ]
```

Refit the model after removing outliers
```{r}
model_energy <- lm(energy ~ architecture * training_environment, data = filtered_energy_data)
summary(model_energy)
```

Now the diagnostics show that the model is well-fitted without significant outliers and the residuals are normally distributed.
```{r}
sim_res_energy <- simulateResiduals(fittedModel = model_energy)

testResiduals(sim_res_energy)
```

## Significance testing
```{r}
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_energy, type = 3)
```

## Post hoc tests

```{r}
emm_energy <- emmeans(model_energy, ~ architecture * training_environment)
pairs(emm_energy, simple = "each", combine = TRUE)
summary(emm_energy)
```

Plot the interaction effects
```{r}
emm_energy_df <- as.data.frame(emm_energy)
emm_df <- emm_energy_df %>%
  mutate(
    energy_estimate = emmean,
    lower_CI = lower.CL,
    upper_CI = upper.CL
  )

# Set the desired order for the x axis
emm_df$architecture <- factor(emm_df$architecture, levels = c("Mobilenet V2", "NASNet Mobile", "ResNet50", "Xception", "VGG16"))

ggplot() +
  # Raw data jittered
  geom_jitter(
    data = filtered_energy_data,
    aes(x = architecture, y = energy, color = training_environment),
    width = 0.2, alpha = 0.3, size = 1
  ) +
  # Estimated points
  geom_point(
    data = emm_df,
    aes(x = architecture, y = energy_estimate, color = training_environment, group = training_environment),
    size = 3, position = position_dodge(width = 0.4)
  ) +
  # Lines connecting estimated points
  geom_line(
    data = emm_df,
    aes(x = architecture, y = energy_estimate, color = training_environment, group = training_environment),
    position = position_dodge(width = 0.4)
  ) +
  # Error bars
  geom_errorbar(
    data = emm_df,
    aes(x = architecture, ymin = lower_CI, ymax = upper_CI, color = training_environment, group = training_environment),
    width = 0.2, position = position_dodge(width = 0.4)
  ) +
  labs(
    title = NULL,
    x = NULL,
    y = "Estimated Energy Consumption (J/image)",
    color = "Training Environment"
  ) +
  theme_pubr() +
  theme(text = element_text(size = 14))
```
---

# GPU Usage Analysis

## Data summary

```{r}
# Outlier filtering for gpu_usage (by group)
filtered_gpu_data <- filter_outliers(df, "gpu_usage")

filtered_gpu_data %>%
  group_by(architecture, training_environment) %>%
  summarise(
    mean_gpu_usage = mean(gpu_usage, na.rm = TRUE),
    sd_gpu_usage = sd(gpu_usage, na.rm = TRUE),
    min_gpu_usage = min(gpu_usage, na.rm = TRUE),
    max_gpu_usage = max(gpu_usage, na.rm = TRUE),
    n = n()
  )
```

```{r}
par(mfrow = c(2, 3)) # Set up a 2x3 grid for histograms
hist(filtered_gpu_data$gpu_usage, main = "Raw")
hist(log(filtered_gpu_data$gpu_usage), main = "Log")
hist(sqrt(filtered_gpu_data$gpu_usage), main = "Sqrt")
hist(filtered_gpu_data$gpu_usage^2, main = "Square")
hist(filtered_gpu_data$gpu_usage^3, main = "Cube")
hist(100 - filtered_gpu_data$gpu_usage, main = "Reflected")
```

## Linear Model
```{r}
max_val <- 100
filtered_gpu_data$gpu_usage_reflected <- sqrt(max_val - filtered_gpu_data$gpu_usage + 1)

model_gpu <- lm(gpu_usage_reflected ~ architecture * training_environment, data = filtered_gpu_data)
summary(model_gpu)
```

## Model diagnostics

```{r}
sim_res_gpu <- simulateResiduals(fittedModel = model_gpu)
testResiduals(sim_res_gpu)
```

### Remove the outliers detected in the residuals and refit the model
The results of the outlier test indicate that there are some outliers in the residuals. We will remove these outliers based on the scaled residuals, which are the standardized residuals from the model.
We will filter out any observations with scaled residuals outside the 2.5%–97.5% quantiles.
```{r}
# Remove outliers based on the simulated residuals
scaled_res_gpu <- sim_res_gpu$scaledResiduals
# Identify outliers: residuals outside the 2.5%–97.5% quantiles
lower_gpu <- quantile(scaled_res_gpu, 0.025)
upper_gpu <- quantile(scaled_res_gpu, 0.975)
# Mark which residuals are outside the acceptable range
outlier_index_gpu <- which(scaled_res_gpu < lower_gpu | scaled_res_gpu > upper_gpu)
filtered_gpu_data <- filtered_gpu_data[-outlier_index_gpu, ]
# Show the outliers
filtered_gpu_data[outlier_index_gpu, ]
```

Refit the model after removing outliers
```{r}
model_gpu <- lm(gpu_usage_reflected ~ architecture * training_environment, data = filtered_gpu_data)
summary(model_gpu)
```

Now the diagnostics show that the model is well-fitted without significant outliers and the residuals are normally distributed.
```{r}
sim_res_gpu <- simulateResiduals(fittedModel = model_gpu)
testResiduals(sim_res_gpu)
```

```{r}
plot(sim_res_gpu)  # DHARMa residual plot
plot(residuals(model_gpu) ~ fitted(model_gpu))  # classic residuals
```

## Significance testing
```{r}
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_gpu, type = 3)
```

## Post hoc tests

```{r}
emmeans(model_gpu, pairwise ~ architecture * training_environment)
```

```{r}
emm_gpu <- emmeans(model_gpu, ~ architecture * training_environment)

# back-transform estimated means
emm_gpu_df <- as.data.frame(emm_gpu)
emm_gpu_df <- emm_gpu_df %>%
  mutate(
    gpu_estimate = max_val - emmean^2 + 1,
    lower_CI = max_val - (emmean + SE * qt(0.975, df))^2 + 1,
    upper_CI = max_val - (emmean - SE * qt(0.975, df))^2 + 1
  )
```

Plot the interaction effects
```{r}
# Set the desired order for the x axis
emm_gpu_df$architecture <- factor(emm_gpu_df$architecture, levels = c("Mobilenet V2", "NASNet Mobile", "ResNet50", "Xception", "VGG16"))

ggplot() +
  # Raw data jittered
  geom_jitter(
    data = filtered_gpu_data,
    aes(x = architecture, y = gpu_usage, color = training_environment),
    width = 0.2, alpha = 0.3, size = 1
  ) +
  # Estimated points
  geom_point(
    data = emm_gpu_df,
    aes(x = architecture, y = gpu_estimate, color = training_environment, group = training_environment),
    size = 3, position = position_dodge(width = 0.4)
  ) +
  # Line connecting the points
  geom_line(
    data = emm_gpu_df,
    aes(x = architecture, y = gpu_estimate, color = training_environment, group = training_environment),
    position = position_dodge(width = 0.4)
  ) +
  # Error bars
  geom_errorbar(
    data = emm_gpu_df,
    aes(x = architecture, ymin = lower_CI, ymax = upper_CI, color = training_environment, group = training_environment),
    width = 0.2, position = position_dodge(width = 0.4)
  ) +
  labs(
    title = NULL,
    x = NULL,
    y = "Estimated GPU Usage (%)",
    color = "Training Environment"
  ) +
  theme_pubr() +
  theme(text = element_text(size = 14),
      axis.text.x = element_text(size = 12))
```

## Visualize estimated energy consumption and gpu usage by architecture and training environment
Show the estimated energy consumption and GPU usage by architecture and training environment in two plots side by side.
```{r fig.align="center", fig.width=14}
energy_plot <- ggplot() +
  geom_jitter(
    data = filtered_energy_data,
    aes(x = architecture, y = energy, color = training_environment),
    width = 0.2, alpha = 0.3, size = 1
  ) +
  geom_point(
    data = emm_df,
    aes(x = architecture, y = energy_estimate, color = training_environment, group = training_environment),
    size = 3, position = position_dodge(width = 0.4)
  ) +
  geom_line(
    data = emm_df,
    aes(x = architecture, y = energy_estimate, color = training_environment, group = training_environment),
    position = position_dodge(width = 0.4)
  ) +
  geom_errorbar(
    data = emm_df,
    aes(x = architecture, ymin = lower_CI, ymax = upper_CI, color = training_environment, group = training_environment),
    width = 0.2, position = position_dodge(width = 0.4)
  ) +
  labs(
    title = NULL,
    x = NULL,
    y = "Estimated Energy Consumption (J/image)",
    color = "Training Environment"
  ) +
  theme_pubr() +
  theme(text = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1))
# Create the GPU usage plot
gpu_plot <- ggplot() +
  geom_jitter(
    data = filtered_gpu_data,
    aes(x = architecture, y = gpu_usage, color = training_environment),
    width = 0.2, alpha = 0.3, size = 1
  ) +
  geom_point(
    data = emm_gpu_df,
    aes(x = architecture, y = gpu_estimate, color = training_environment, group = training_environment),
    size = 3, position = position_dodge(width = 0.4)
  ) +
  geom_line(
    data = emm_gpu_df,
    aes(x = architecture, y = gpu_estimate, color = training_environment, group = training_environment),
    position = position_dodge(width = 0.4)
  ) +
  geom_errorbar(
    data = emm_gpu_df,
    aes(x = architecture, ymin = lower_CI, ymax = upper_CI, color = training_environment, group = training_environment),
    width = 0.2, position = position_dodge(width = 0.4)
  ) +
  labs(
    title = NULL,
    x = NULL,
    y = "Estimated GPU Usage (%)",
    color = "Training Environment"
  ) +
  theme_pubr() +
  theme(text = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1))
# grid.arrange(energy_plot, gpu_plot, ncol = 2)

combined_plot <- ggarrange(
  energy_plot, gpu_plot,
  ncol = 2, nrow = 1,
  common.legend = TRUE,
  legend = "top"
)
combined_plot
```
