---
title: "RQ1: What is the impact of model architecture and training environment on image classification training concerning energy consumption and GPU usage?"
output: html_notebook
author: "Santiago del Rey"
---

```{r include=FALSE}
# Load libraries
library(arrow)
library(dplyr)
library(lme4)
library(lmerTest) # p-values for lmer
library(DHARMa)
library(emmeans) # post hoc tests
library(ggplot2)
library(ggpubr) # for publication-ready plots
library(car)
library(MASS)
```

## Load and preprocess data

We load the data from the parquet file, filter out the `inception_v3` architecture and the `Desktop Normal User` training environment, and normalize the energy consumption by the total number of images seen during training (converting MJ to J). We also prepare the `gpu_usage` variable.

```{r}
df <- read_parquet("../data/metrics/processed/dl-training-energy-consumption-dataset.gzip") %>%
  filter(architecture != "inception_v3" & `training environment` != "Desktop Normal User" & `return code` == 0) %>%
  mutate(
    energy = (`energy (MJ)` * 1000000) / `total seen images`, # Convert MJ to J and normalize by total seen images
    training_environment = factor(`training environment`),
    architecture = factor(architecture),
    gpu_usage = `gpu usage (%)`
  ) %>%
  dplyr::select(energy, gpu_usage, architecture, training_environment)
```

### Outlier filtering functions

```{r}
# Define a function to filter outliers of a given variable on an input data frame
filter_outliers <- function(df, variable) {
  df %>%
    group_by(architecture, training_environment) %>%
    mutate(
        Q1 = quantile(!!sym(variable), 0.25),
        Q3 = quantile(!!sym(variable), 0.75),
        IQR = Q3 - Q1,
        is_outlier = !!sym(variable) < (Q1 - 1.5 * IQR) | !!sym(variable) > (Q3 + 1.5 * IQR)
    ) %>%
    ungroup() %>%
    filter(!is_outlier) %>%
    dplyr::select(-Q1, -Q3, -IQR, -is_outlier)
}
```

---

# Energy Consumption Analysis

## Data summary

```{r}
filtered_energy_data <- df %>%
  filter_outliers("energy")

filtered_energy_data %>%
  group_by(architecture, training_environment) %>%
  summarise(
    mean_energy = mean(energy, na.rm = TRUE),
    sd_energy = sd(energy, na.rm = TRUE),
    min_energy = min(energy, na.rm = TRUE),
    max_energy = max(energy, na.rm = TRUE),
    n = n()
  )
```

```{r}
hist(filtered_energy_data$energy, breaks = 30, main = "Distribution of Energy")
```

## Linear Model

```{r}
model_energy <- lm(energy ~ architecture * training_environment, data = filtered_energy_data)
summary(model_energy)
```

## Model diagnostics

```{r}
sim_res_energy <- simulateResiduals(fittedModel = model_energy)
testDispersion(sim_res_energy)        # Should not be significant
testUniformity(sim_res_energy)        # Check for deviation from expected residuals
testOutliers(sim_res_energy)          # Identify outlier behavior
```

## Significance testing

```{r}
options(contrasts = c("contr.sum", "contr.poly"))  # sum-to-zero contrasts for Type III
Anova(model_energy, type = 3)
```

## Post hoc tests

```{r}
emm_energy <- emmeans(model_energy, ~ architecture * training_environment)
pairs(emm_energy, simple = "each", combine = TRUE)
summary(emm_energy)
```

---

# GPU Usage Analysis

## Data summary

```{r}
# Outlier filtering for gpu_usage (by group)
filtered_gpu_data <- filter_outliers(df, "gpu_usage")

filtered_gpu_data %>%
  group_by(architecture, training_environment) %>%
  summarise(
    mean_gpu_usage = mean(gpu_usage, na.rm = TRUE),
    sd_gpu_usage = sd(gpu_usage, na.rm = TRUE),
    min_gpu_usage = min(gpu_usage, na.rm = TRUE),
    max_gpu_usage = max(gpu_usage, na.rm = TRUE),
    n = n()
  )
```

```{r}
hist(filtered_gpu_data$gpu_usage, breaks = 30, main = "Distribution of GPU usage")
```

```{r}
boxcox(lm(gpu_usage ~ architecture * training_environment, data = filtered_gpu_data))
```

## Linear Model

```{r}
filtered_gpu_data$gpu_usage_sq <- filtered_gpu_data$gpu_usage^2
model_gpu <- lm(gpu_usage_sq ~ architecture * training_environment, data = filtered_gpu_data)
summary(model_gpu)
```

## Model diagnostics

```{r}
sim_res_gpu <- simulateResiduals(fittedModel = model_gpu)
testDispersion(sim_res_gpu)        # Should not be significant
testUniformity(sim_res_gpu)        # Check for deviation from expected residuals
testOutliers(sim_res_gpu)          # Identify outlier behavior
```

## Significance testing

```{r}
options(contrasts = c("contr.sum", "contr.poly"))  # sum-to-zero contrasts for Type III
Anova(model_gpu, type = 3)
```

## Post hoc tests

```{r}
emm_gpu <- emmeans(model_gpu, ~ architecture * training_environment)
pairs(emm_gpu, simple = "each", combine = TRUE)
summary(emm_gpu)
```