---
title: "RQ1 and RQ2"
output: html_notebook
author: "Santiago del Rey"
---

## Load libraries
```{r include=FALSE}
library(arrow)
library(dplyr)
library(lme4)
library(lmerTest)      # p-values for lmer
library(performance)   # check assumptions
library(emmeans)       # post hoc tests
library(car)
library(ARTool)
library(ggplot2)

```

## Load data
We load the data from the parquet file. We filter out the `inception_v3` architecture and the `Local Normal User` training environment, as they are not relevant for our analysis. We also rename the `energy (MJ)` column to `energy` for convenience.
```{r}
df <- read_parquet("../data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
  filter(architecture != "inception_v3" & `training environment` != "Local Normal User") %>%
  mutate(
    energy=`energy (MJ)`,
    training_environment= factor(`training environment`),
    base_model=factor(architecture),
    ) %>%
    group_by(base_model, training_environment) %>%
    mutate(
      subject_id = factor(row_number()),
    )
```

## Create a linear mixed model
We create a linear mixed model with `energy` as the response variable and `training_environment` and `base_model` as fixed effects. We also include an error term for the `subject` variable to account for the fact that it is a random effect.
```{r}
model_lmm <- lmer(energy ~ base_model * training_environment + (1 | subject_id), data = df)
summary(model_lmm)
```

## Check assumptions
We check the assumptions of the linear mixed model using the `performance` package. We check for normality of residuals, homoscedasticity, and independence of residuals.
```{r}
# QQ plot
qqnorm(resid(model_lmm)); qqline(resid(model_lmm))

# Residuals histogram
hist(resid(model_lmm), main = "Residuals Histogram", xlab = "Residuals")

# Residuals vs Fitted
plot(fitted(model_lmm), resid(model_lmm), main = "Residuals vs Fitted", xlab = "Fitted", ylab = "Residuals")
abline(h = 0, col = "red")
```

## Test normality of residuals
We use the Shapiro-Wilk test to check the normality of the residuals.
```{r}
normality_results <- shapiro.test(resid(model_lmm))  # If p < 0.05, consider transformation
normality_results
```

```{r}
if (normality_results$p.value < 0.05) {
  cat("Residuals are not normally distributed. Consider transformation.\n")
}
```

## Fit a GLMM
We fit a generalized linear mixed model (GLMM) with a Gamma distribution and a log link function. This is appropriate for positive continuous data like energy consumption.
```{r}
model_glmm <- glmer(energy ~ base_model * training_environment + (1 | subject_id), data = df, family = Gamma(link = "log"), control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))
summary(model_glmm)
```

Exponentiated coefficients of the GLMM give us the multiplicative effect of each predictor on the response variable.
```{r}
exp(fixef(model_glmm))
```


## Post hoc tests
We use the `emmeans` package to perform post hoc tests on the linear mixed model. We test for differences in energy consumption between different training environments and base models.
```{r}
emm <- emmeans(model_glmm, ~ base_model * training_environment)
plot(emm, comparisons = TRUE) +
  theme_minimal()
```


Make pairwise comparisons between the estimated marginal means. We use Tukey's method for multiple comparisons.
```{r}
# Pairwise comparisons
pairs(emm, simple = "base_model", adjust = "tukey")
pairs(emm, simple = "training_environment", adjust = "tukey")
```
```{r}
pwpp(emm, by = "training_environment", type = "response")
```


## Check for overdispersion
We check for overdispersion in the GLMM using the Pearson chi-squared statistic. If the ratio of the Pearson chi-squared statistic to the residual degrees of freedom is greater than 1.5, we have overdispersion.
```{r}
overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp <- residuals(model, type = "pearson")
  Pearson.chisq <- sum(rp^2)
  prat <- Pearson.chisq / rdf
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
  c(chisq = Pearson.chisq, ratio = prat, rdf = rdf, p = pval)
}
overdisp_fun(model_glmm)
```
