---
title: "RQ1: What is the impact of model architecture and training environment on image classification training concerning energy consumption and GPU usage?"
output: html_notebook
author: "Santiago del Rey"
---

```{r include=FALSE}
# Load libraries
library(arrow)
library(dplyr)
library(lme4)
library(lmerTest) # p-values for lmer
library(performance) # check assumptions
library(DHARMa)
library(glmmTMB)
library(lattice)
library(emmeans) # post hoc tests
library(MASS)
library(car)
library(minpack.lm)
library(boot)
library(sjPlot)
library(ggplot2)
library(ggpubr) # for publication-ready plots
```

## Load data
We load the data from the parquet file. We filter out the `inception_v3` architecture and the `Local Normal User` training environment, as they are not relevant for RQ1.
We also normalize the energy consumption by the total number of images seen during training and convert the energy from MJ to J for analysis.
```{r}
df <- read_parquet("../data/metrics/processed/dl-training-energy-consumption-dataset.gzip") %>%
  filter(architecture != "inception_v3" & `training environment` != "Desktop Normal User" & `return code` == 0) %>%
  mutate(
    energy = (`energy (MJ)` * 1000000) / `total seen images`, # Convert MJ to J and normalize by total seen images
    training_environment = factor(`training environment`),
    architecture = factor(architecture),
    gpu_usage = `gpu usage (%)`,
  )

clean_data <- df %>%
  group_by(architecture, training_environment) %>%
  mutate(
    Q1 = quantile(energy, 0.25),
    Q3 = quantile(energy, 0.75),
    IQR = Q3 - Q1,
    is_outlier = energy < (Q1 - 1.5 * IQR) | energy > (Q3 + 1.5 * IQR)
  ) %>%
  ungroup()

filtered_data <- clean_data %>%
  filter(!is_outlier) %>%
  dplyr::select(-Q1, -Q3, -IQR, -is_outlier)
```

```{r}
head(df)
```

## Data summary
```{r}
filtered_data %>%
  group_by(architecture, training_environment) %>%
  summarise(
    mean_energy = mean(energy, na.rm = TRUE),
    sd_energy = sd(energy, na.rm = TRUE),
    min_energy = min(energy, na.rm = TRUE),
    max_energy = max(energy, na.rm = TRUE),
    n = n()
  )
```

```{r}
hist(filtered_data$energy, breaks = 30, main = "Distribution of Energy")
```


## Create a linear mixed model
We create a linear mixed model with `energy` as the response variable and `training_environment` and `architecture` as fixed effects. We also include an error term for the `subject` variable to account for the fact that it is a random effect.
```{r}
# df$log_energy <- log(df$energy)
model <- lm(energy ~ architecture * training_environment, data = filtered_data)

summary(model)
```

## Model diagnostics
We check the model diagnostics to ensure that the LMM is appropriate for the data.
```{r}
sim_res <- simulateResiduals(fittedModel = model)

testDispersion(sim_res)        # Should not be significant
testUniformity(sim_res)        # Check for deviation from expected residuals
testOutliers(sim_res)          # Identify outlier behavior
```

## Significance testing
```{r}
# For linear models
options(contrasts = c("contr.sum", "contr.poly"))  # sum-to-zero contrasts for Type III
Anova(model, type = 3)
```

## Post hoc tests
```{r}
emm <- emmeans(model, ~ architecture * training_environment)

# Pairwise comparisons
pairs(emm, simple = "each", combine = TRUE)

# Summary of marginal means
summary(emm)
```
