---
title: "RQ2: Do gains in correctness justify energy-greedy architectures?"
output: html_notebook
author: "Santiago del Rey"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
library(arrow)
library(tidyverse)
library(lme4)
library(lmerTest) # p-values for lmer
library(betareg)
library(DHARMa)
library(broom)
library(car)
library(performance) # check assumptions
library(emmeans) # post hoc tests
library(multcomp)
library(gridExtra)
library(ggfortify)
library(ggpubr) # for publication-ready plots
```

## Load and preprocess data

We load the data from the parquet file, filter out the `inception_v3` architecture and the `Desktop Normal User` training environment, and normalize the energy consumption by the total number of images seen during training (converting MJ to J). We also prepare the `gpu_usage` variable.

```{r}
df <- read_parquet("../data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
    rename(training_environment = `training environment`, gpu_usage = `gpu usage (%)`, f1_score = `f1-score`) %>%
    filter(architecture != "inception_v3" & training_environment != "Desktop Normal User") %>%
    mutate(
        energy_per_image = (`energy (MJ)` * 1000000) / `total seen images`, # Convert MJ to J and normalize by total seen images
        architecture = case_match(
            architecture,
            "mobilenet_v2" ~ "Mobilenet V2",
            "nasnet_mobile" ~ "NASNet Mobile",
            "resnet50" ~ "ResNet50",
            "xception" ~ "Xception",
            "vgg16" ~ "VGG16"
        ),
        architecture = factor(architecture, levels = c("Mobilenet V2", "NASNet Mobile", "ResNet50", "Xception", "VGG16")),
        training_environment = factor(training_environment, levels = c("Desktop ML Engineer", "Server")),
        efficiency = f1_score / energy_per_image,
    ) %>%
    mutate(energy_per_image = as.numeric(scale(energy_per_image, center = TRUE, scale = TRUE))) %>%
    dplyr::select(energy_per_image, f1_score, efficiency, GFLOPs, architecture, training_environment)
```

### Outlier filtering functions

```{r}
# Define a function to filter outliers of a given variable on an input data frame
filter_outliers <- function(df, variable) {
    df %>%
        group_by(architecture, training_environment) %>%
        mutate(
            Q1 = quantile(!!sym(variable), 0.25),
            Q3 = quantile(!!sym(variable), 0.75),
            IQR = Q3 - Q1,
            is_outlier = !!sym(variable) < (Q1 - 1.5 * IQR) | !!sym(variable) > (Q3 + 1.5 * IQR)
        ) %>%
        ungroup() %>%
        filter(!is_outlier) %>%
        dplyr::select(-Q1, -Q3, -IQR, -is_outlier)
}
```

## Data overview
```{r data_overview}
# Filter F1-score outliers
df_filtered_f1 <- df %>%
    group_by(architecture) %>%
    mutate(
        Q1 = quantile(f1_score, 0.25),
        Q3 = quantile(f1_score, 0.75),
        IQR = Q3 - Q1,
        is_outlier = f1_score < (Q1 - 1.5 * IQR) | f1_score > (Q3 + 1.5 * IQR)
    ) %>%
    ungroup() %>%
    filter(!is_outlier) %>%
    dplyr::select(-Q1, -Q3, -IQR, -is_outlier)


print(paste("Number of outliers removed from f1_score:", nrow(df) - nrow(df_filtered_f1)))

df_clean <- filter_outliers(df_filtered_f1, "energy_per_image")
print(paste("Number of outliers removed from energy_per_image:", nrow(df_filtered_f1) - nrow(df_clean)))
# Display the first few rows of the data frame
head(df)
# Display summary statistics of the data frame
summary(df)

df %>%
    group_by(architecture, training_environment) %>%
    summarise(
        min_f1_score = min(f1_score, na.rm = TRUE),
        max_f1_score = max(f1_score, na.rm = TRUE),
        mean_f1_score = mean(f1_score, na.rm = TRUE),
        sd_f1_score = sd(f1_score, na.rm = TRUE),
        min_energy_per_image = min(energy_per_image, na.rm = TRUE),
        max_energy_per_image = max(energy_per_image, na.rm = TRUE),
        mean_energy_per_image = mean(energy_per_image, na.rm = TRUE),
        sd_energy_per_image = sd(energy_per_image, na.rm = TRUE),
        min_efficiency = min(efficiency, na.rm = TRUE),
        max_efficiency = max(efficiency, na.rm = TRUE),
        mean_efficiency = mean(efficiency, na.rm = TRUE),
        sd_efficiency = sd(efficiency, na.rm = TRUE),
    ) %>%
    ungroup() %>%
    arrange(architecture, training_environment)
```

## Exploratory visualization
### Visualize the GFLOPs of each architecture
```{r gflops_by_arch, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
ggplot(df, aes(x = architecture, y = GFLOPs)) +
    geom_point() +
    labs(title = "GFLOPs across Architectures") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
```

### Visualize energy_per_image vs. F1-score, colored by architecture
```{r energy_vs_f1, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
ggscatter(df_clean,
    x = "energy_per_image", y = "f1_score",
    color = "architecture", facet.by = "training_environment",
    add = "reg.line", conf.int = TRUE,
    palette = "Set1",
) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(title = "energy_per_image vs. F1-score by Architecture and Training Environment")
```

### Efficiency: F1-score per Megajoule
```{r f1_per_joule, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
ggbarplot(df_clean,
    x = "architecture", y = "efficiency",
    fill = "training_environment", position = position_dodge(),
    add = "mean_se", palette = "Set1"
) +
    labs(title = "Efficiency (F1-score per Megajoule) by Architecture and Training Environment")
```


### Visualize histograms for F1 score
```{r f1_histograms}
par(mfrow = c(2, 3)) # Set up a 2x3 grid for histograms
hist(df_clean$f1_score, main = "Raw")
hist(log(df_clean$f1_score), main = "Log")
hist(sqrt(df_clean$f1_score), main = "Sqrt")
hist(df_clean$f1_score^2, main = "Square")
hist(df_clean$f1_score^3, main = "Cube")
hist(log(df$f1_score / (1 - df$f1_score)), main = "Logit")
```

## Statistical Analysis
### Model 1: F1-score as a function of architecture and training environment
```{r model1}
df_clean$f1_logit <- log(df_clean$f1_score / (1 - df_clean$f1_score))
model1 <- lm(f1_logit ~ architecture * training_environment, data = df_clean)
summary(model1)
```
### Model diagnostics for Model 1
```{r model1_diagnostics}
# Check residuals
residuals_model1 <- residuals(model1)
sim_res_model1 <- simulateResiduals(model1)
# Plot residuals
testResiduals(sim_res_model1)
# Check homoscedasticity
homoscedasticity_test <- leveneTest(residuals_model1 ~ architecture * training_environment, data = df_clean)
homoscedasticity_test
```
```{r check_model1, fig.height=6}
# Check model assumptions
model1_performance <- performance::check_model(model1)
model1_performance
```
```{r}
# Check for outliers
outliers_model1 <- model1_performance$outliers
outliers_model1
# Post hoc tests
post_hoc_model1 <- emmeans(model1, ~ architecture * training_environment)
post_hoc_results_model1 <- summary(post_hoc_model1)
post_hoc_results_model1
# Plot post hoc results
post_hoc_plot_model1 <- ggplot(post_hoc_results_model1, aes(x = architecture, y = emmean, fill = training_environment)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, position = position_dodge(0.9)) +
    labs(title = "Post Hoc Results for Model 1", x = "Architecture", y = "Estimated Marginal Mean (F1-score)") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
print(post_hoc_plot_model1)
```

### Remove the outliers detected in the residuals and refit the model
The results of the outlier test indicate that there are some outliers in the residuals. We will remove these outliers based on the scaled residuals, which are the standardized residuals from the model.
We will filter out any observations with scaled residuals outside the 2.5%–97.5% quantiles.
```{r}
# Remove outliers based on the simulated residuals
scaled_res_model1 <- sim_res_model1$scaledResiduals
# Identify outliers: residuals outside the 2.5%–97.5% quantiles
lower_f1_logit <- quantile(scaled_res_model1, 0.025)
upper_f1_logit <- quantile(scaled_res_model1, 0.975)
# Mark which residuals are outside the acceptable range
outlier_index_model1 <- which(scaled_res_model1 < lower_f1_logit | scaled_res_model1 > upper_f1_logit)
df_filtered_model1 <- df_clean[-outlier_index_model1, ]
# Show the outliers
df_filtered_model1[outlier_index_model1, ]
```

### Refit Model 1 after removing residual outliers
```{r model1_filtered}
model1_filtered <- lm(f1_logit ~ architecture * training_environment, data = df_filtered_model1)
summary(model1_filtered)
```
```{r model1_diagnostics2}
# Check residuals
residuals_model1 <- residuals(model1_filtered)
sim_res_model1 <- simulateResiduals(model1_filtered)
# Plot residuals
testResiduals(sim_res_model1)
```

### Model 2: F1-score per Joule as a function of architecture and training environment
```{r model2}
model2 <- lm(efficiency ~ architecture * training_environment, data = df_clean)
summary(model2)
```
### Model diagnostics for Model 2
```{r model2_diagnostics}
# Check residuals
sim_res_model2 <- simulateResiduals(model2)
# Plot residuals
testResiduals(sim_res_model2)
# Check homoscedasticity
homoscedasticity_test_model2 <- leveneTest(residuals(model2) ~ architecture * training_environment, data = df_clean)
homoscedasticity_test_model2
```
```{r fig.height=10}
# Check model assumptions
model2_performance <- performance::check_model(model2)
model2_performance
```
```{r}
# Check for outliers
outliers_model2 <- model2_performance$outliers
outliers_model2
# Post hoc tests
post_hoc_model2 <- emmeans(model2, ~ architecture * training_environment)
post_hoc_results_model2 <- summary(post_hoc_model2)
post_hoc_results_model2
# Plot post hoc results
post_hoc_plot_model2 <- ggplot(post_hoc_results_model2, aes(x = architecture, y = emmean, fill = training_environment)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, position = position_dodge(0.9)) +
    labs(title = "Post Hoc Results for Model 2", x = "Architecture", y = "Estimated Marginal Mean (F1-score per Joule)") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
print(post_hoc_plot_model2)
```

### Model 3: F1-score as a function of energy_per_image, architecture, and training environment
```{r model3}
model3 <- lm(f1_logit ~ energy_per_image + architecture * training_environment, data = df_clean)
summary(model3)
```
### Model diagnostics for Model 3
```{r model3_diagnostics}
# Check residuals
residuals_model3 <- residuals(model3)
sim_res_model3 <- simulateResiduals(model3)
# Plot residuals
testResiduals(sim_res_model3)
# Check homoscedasticity
# homoscedasticity_test_model3 <- leveneTest(residuals_model3 ~ energy_per_image + architecture * training_environment, data = df)
# homoscedasticity_test_model3
```
```{r fig.height=6}
# Check model assumptions
model3_performance <- performance::check_model(model3)
model3_performance
```
```{r}
# Check for outliers
outliers_model3 <- model3_performance$outliers
outliers_model3
# Post hoc tests
post_hoc_model3 <- emmeans(model3, ~ architecture * training_environment)
post_hoc_results_model3 <- summary(post_hoc_model3)
# Plot post hoc results
post_hoc_plot_model3 <- ggplot(post_hoc_results_model3, aes(x = architecture, y = emmean, fill = training_environment)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, position = position_dodge(0.9)) +
    labs(title = "Post Hoc Results for Model 3", x = "Architecture", y = "Estimated Marginal Mean (F1-score)") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
print(post_hoc_plot_model3)
```

### Model 3B: Use beta regression
```{r model3b}
df_clean <- df_clean %>% mutate(f1_trans = (f1_score * (n() - 1) + 0.5) / n())
model3b <- betareg(f1_trans ~ architecture * training_environment | energy_per_image, data = df_clean)
summary(model3b)
```

```{r}
summary(lm(f1_score ~ energy_per_image, data = df_clean))
```


### Model diagnostics for Model 1B
```{r model3b_diagnostics}
# Check residuals
residuals_model3b <- residuals(model3b)
# Plot residuals
par(mfrow = c(2, 2))
plot(model3b)
```

Show extreme residuals
```{r extreme_residuals}
extreme_residuals <- which(abs(residuals(model3b, type = "pearson")) > 2)

df_clean[extreme_residuals, ]
```
```{r model3b_post_hoc}
# Post hoc tests
post_hoc_model3b <- emmeans(model3b, ~ architecture * training_environment, type = "response")
post_hoc_results_model3b <- summary(post_hoc_model3b)
post_hoc_results_model3b
pairs(post_hoc_model3b)
# Plot post hoc results
post_hoc_plot_model3b <- ggplot(post_hoc_results_model3b, aes(x = architecture, y = emmean, fill = training_environment)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.2, position = position_dodge(0.9)) +
    labs(title = "Post Hoc Results for Model 1", x = "Architecture", y = "Estimated Marginal Mean (F1-score)") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
print(post_hoc_plot_model3b)

cld(post_hoc_model3b, Letters = letters, adjust = "tukey")

```


## Model Comparison
```{r model_comparison}
# Compare models using ANOVA
anova(model1, model2, model3)
# Compare models using AIC
model_aic <- AIC(model1, model2, model3)
model_aic <- model_aic %>%
    mutate(model = c("Model 1", "Model 2", "Model 3")) %>%
    arrange(AIC)
# Plot AIC comparison
ggplot(model_aic, aes(x = model, y = AIC)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(title = "AIC Comparison of Models", x = "Model", y = "AIC") +
    theme_minimal()
# Compare models using BIC
model_bic <- BIC(model1, model2, model3)
model_bic <- model_bic %>%
    mutate(model = c("Model 1", "Model 2", "Model 3")) %>%
    arrange(BIC)
# Plot BIC comparison
ggplot(model_bic, aes(x = model, y = BIC)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(title = "BIC Comparison of Models", x = "Model", y = "BIC") +
    theme_minimal()
# Compare models using R-squared
model_r2 <- data.frame(
    model = c("Model 1", "Model 2", "Model 3"),
    r_squared = c(performance::r2(model1), performance::r2(model2), performance::r2(model3))
)
# ggplot(model_r2, aes(x = model, y = r_squared)) +
#     geom_bar(stat = "identity", fill = "steelblue") +
#     labs(title = "R-squared Comparison of Models", x = "Model", y = "R-squared") +
#     theme_minimal()
# Compare models using adjusted R-squared
model_adj_r2 <- data.frame(
    model = c("Model 1", "Model 2", "Model 3"),
    adj_r_squared = c(performance::r2(model1), performance::r2(model2, adjusted = TRUE), performance::r2(model3, adjusted = TRUE))
)
# ggplot(model_adj_r2, aes(x = model, y = adj_r_squared)) +
#     geom_bar(stat = "identity", fill = "steelblue") +
#     labs(title = "Adjusted R-squared Comparison of Models", x = "Model", y = "Adjusted R-squared") +
#     theme_minimal()
# Compare models using RMSE
model_rmse <- data.frame(
    model = c("Model 1", "Model 2", "Model 3"),
    rmse = c(performance::rmse(model1), performance::rmse(model2), performance::rmse(model3))
)
ggplot(model_rmse, aes(x = model, y = rmse)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(title = "RMSE Comparison of Models", x = "Model", y = "RMSE") +
    theme_minimal()
# Compare models using MAE
model_mae <- data.frame(
    model = c("Model 1", "Model 2", "Model 3"),
    mae = c(performance::mae(model1), performance::mae(model2), performance::mae(model3))
)
ggplot(model_mae, aes(x = model, y = mae)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(title = "MAE Comparison of Models", x = "Model", y = "MAE") +
    theme_minimal()
# Compare models using MSE
model_mse <- data.frame(
    model = c("Model 1", "Model 2", "Model 3"),
    mse = c(performance::mse(model1), performance::mse(model2), performance::mse(model3))
)
ggplot(model_mse, aes(x = model, y = mse)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(title = "MSE Comparison of Models", x = "Model", y = "MSE") +
    theme_minimal()
```