---
title: "RQ2: How does the computational complexity of model architectures affect the training energy consumption, GPU usage, and model correctness?"
output: html_notebook
author: "Santiago del Rey"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
library(arrow)
library(tidyverse)
library(lme4)
library(lmerTest) # p-values for lmer
library(DHARMa)
library(broom)
library(car)
library(performance) # check assumptions
library(emmeans) # post hoc tests
library(gridExtra)
library(ggfortify)
library(ggpubr) # for publication-ready plots
```

## Load and preprocess data

We load the data from the parquet file, filter out the `inception_v3` architecture and the `Desktop Normal User` training environment, and normalize the energy consumption by the total number of images seen during training (converting MJ to J). We also prepare the `gpu_usage` variable.

```{r}
df <- read_parquet("../data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
    rename(training_environment = `training environment`, gpu_usage = `gpu usage (%)`, f1_score = `f1-score`) %>%
    filter(architecture != "inception_v3" & training_environment != "Desktop Normal User") %>%
    mutate(
        energy = (`energy (MJ)` * 1000000) / `total seen images`, # Convert MJ to J and normalize by total seen images
        architecture = case_match(
            architecture,
            "mobilenet_v2" ~ "Mobilenet V2",
            "nasnet_mobile" ~ "NASNet Mobile",
            "resnet50" ~ "ResNet50",
            "xception" ~ "Xception",
            "vgg16" ~ "VGG16"
        ),
        architecture = factor(architecture, levels = c("Mobilenet V2", "NASNet Mobile", "ResNet50", "Xception", "VGG16")),
        training_environment = factor(training_environment, levels = c("Desktop ML Engineer", "Server")),
    ) %>%
    dplyr::select(energy, gpu_usage, f1_score, GFLOPs, architecture, training_environment)
```

### Outlier filtering functions

```{r}
# Define a function to filter outliers of a given variable on an input data frame
filter_outliers <- function(df, variable) {
    df %>%
        group_by(architecture, training_environment) %>%
        mutate(
            Q1 = quantile(!!sym(variable), 0.25),
            Q3 = quantile(!!sym(variable), 0.75),
            IQR = Q3 - Q1,
            is_outlier = !!sym(variable) < (Q1 - 1.5 * IQR) | !!sym(variable) > (Q3 + 1.5 * IQR)
        ) %>%
        ungroup() %>%
        filter(!is_outlier) %>%
        dplyr::select(-Q1, -Q3, -IQR, -is_outlier)
}
```

## Visualize the GFLOPs
```{r gflops_by_arch, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
ggplot(df, aes(x = architecture, y = GFLOPs)) +
    geom_point() +
    labs(title = "GFLOPs across Architectures") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
```

---

# RQ2.1: How does the architectural computational complexity of a DL model affect energy consumption and GPU usage?

## Effects on energy consumption

Filter outliers for energy consumption and summarize statistics.
```{r}
# Filter outliers for energy consumption
df_energy <- filter_outliers(df, "energy")

# summary statistics for energy consumption
df_energy %>%
    group_by(architecture, training_environment) %>%
    summarise(
        mean_energy = mean(energy),
        sd_energy = sd(energy),
        min_energy = min(energy),
        max_energy = max(energy),
        n = n()
    )
```

```{r, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
par(mfrow = c(2, 3)) # Set up a 2x3 grid for histograms
hist(df_energy$energy, main = "Raw")
hist(log(df_energy$energy), main = "Log")
hist(sqrt(df_energy$energy), main = "Sqrt")
hist(df_energy$energy^2, main = "Square")
hist(df_energy$energy^3, main = "Cube")
hist(100 - df_energy$energy, main = "Reflected")
```

## Initial Linear Model
```{r}

model_energy <- lmer(energy ~ GFLOPs * training_environment + (1 + training_environment | architecture), data = df_energy)
summary(model_energy)
```

### Diagnostics for Initial Model
Base diagnostics
```{r}
par(mfrow = c(2, 2))
plot(model_energy)
```

DHARMa diagnostics
```{r}
sim_energy <- simulateResiduals(model_energy, n = 5000)
plot(sim_energy)
testUniformity(sim_energy)
testDispersion(sim_energy)
testResiduals(sim_energy)
```

```{r}
plotResiduals(sim_energy, df_energy$GFLOPs)
plotResiduals(sim_energy, df_energy$training_environment)
```

```{r}
library(mgcv)
model_gamm <- gamm(energy ~ s(GFLOPs) * training_environment, random = list(architecture = ~1), data = df_energy)
simulationOutput_gamm <- simulateResiduals(model_gamm$mer)
testUniformity(simulationOutput_gamm)
plot(simulationOutput_gamm)
```
