---
title: "RQ2: Do gains in correctness justify energy-greedy architectures?"
output: html_notebook
author: "Santiago del Rey"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
library(arrow)
library(tidyverse)
library(lme4) # for linear mixed-effects models
library(lmerTest) # for p-values in lmer models
library(betareg)
library(glmmTMB) # for beta regression
library(DHARMa)
library(broom)
library(car)
library(performance) # check assumptions
library(emmeans) # post hoc tests
library(multcomp)
library(h2o)
library(gridExtra)
library(ggplot2)
library(ggfortify)
library(ggpubr) # for publication-ready plots
```

## Load and preprocess data

We load the data from the parquet file, filter out the `inception_v3` architecture and the `Desktop Normal User` training environment, and normalize the energy consumption by the total number of images seen during training (converting MJ to J). We also prepare the `gpu_usage` variable.

```{r}
df <- read_parquet("../data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
    rename(training_environment = `training environment`, gpu_usage = `gpu usage (%)`, f1_score = `f1-score`, energy = `energy (MJ)`) %>%
    filter(architecture != "inception_v3" & training_environment != "Desktop Normal User") %>%
    mutate(
        energy_per_image = (energy * 1000000) / `total seen images`, # Convert MJ to J and normalize by total seen images
        architecture = case_match(
            architecture,
            "mobilenet_v2" ~ "Mobilenet V2",
            "nasnet_mobile" ~ "NASNet Mobile",
            "resnet50" ~ "ResNet50",
            "xception" ~ "Xception",
            "vgg16" ~ "VGG16"
        ),
        architecture = factor(architecture, levels = c("Mobilenet V2", "NASNet Mobile", "ResNet50", "Xception", "VGG16")),
        training_environment = factor(training_environment, levels = c("Desktop ML Engineer", "Server")),
        efficiency = energy / f1_score, # Megajoule per F1-score
    ) %>%
    dplyr::select(energy, energy_per_image, f1_score, efficiency, GFLOPs, architecture, training_environment)
```

### Outlier filtering functions

```{r}
# Define a function to filter outliers of a given variable on an input data frame
filter_outliers <- function(df, variable) {
    df %>%
        group_by(architecture, training_environment) %>%
        mutate(
            Q1 = quantile(!!sym(variable), 0.25),
            Q3 = quantile(!!sym(variable), 0.75),
            IQR = Q3 - Q1,
            is_outlier = !!sym(variable) < (Q1 - 1.5 * IQR) | !!sym(variable) > (Q3 + 1.5 * IQR)
        ) %>%
        ungroup() %>%
        filter(!is_outlier) %>%
        dplyr::select(-Q1, -Q3, -IQR, -is_outlier)
}
```

## Data overview
```{r data_overview}
# Filter F1-score outliers
df_filtered_f1 <- df %>%
    group_by(architecture) %>%
    mutate(
        Q1 = quantile(f1_score, 0.25),
        Q3 = quantile(f1_score, 0.75),
        IQR = Q3 - Q1,
        is_outlier = f1_score < (Q1 - 1.5 * IQR) | f1_score > (Q3 + 1.5 * IQR)
    ) %>%
    ungroup() %>%
    filter(!is_outlier) %>%
    dplyr::select(-Q1, -Q3, -IQR, -is_outlier)


print(paste("Number of outliers removed from f1_score:", nrow(df) - nrow(df_filtered_f1)))

df_clean <- filter_outliers(df_filtered_f1, "energy_per_image")
print(paste("Number of outliers removed from energy_per_image:", nrow(df_filtered_f1) - nrow(df_clean)))
# Display the first few rows of the data frame
head(df)
# Display summary statistics of the data frame
summary(df)

df %>%
    group_by(architecture, training_environment) %>%
    summarise(
        min_f1_score = min(f1_score, na.rm = TRUE),
        max_f1_score = max(f1_score, na.rm = TRUE),
        mean_f1_score = mean(f1_score, na.rm = TRUE),
        sd_f1_score = sd(f1_score, na.rm = TRUE),
        min_energy_per_image = min(energy_per_image, na.rm = TRUE),
        max_energy_per_image = max(energy_per_image, na.rm = TRUE),
        mean_energy_per_image = mean(energy_per_image, na.rm = TRUE),
        sd_energy_per_image = sd(energy_per_image, na.rm = TRUE),
        min_efficiency = min(efficiency, na.rm = TRUE),
        max_efficiency = max(efficiency, na.rm = TRUE),
        mean_efficiency = mean(efficiency, na.rm = TRUE),
        sd_efficiency = sd(efficiency, na.rm = TRUE),
    ) %>%
    ungroup() %>%
    arrange(architecture, training_environment)
```

## Exploratory visualization
### Visualize the GFLOPs of each architecture
```{r gflops_by_arch, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
ggplot(df, aes(x = architecture, y = GFLOPs)) +
    geom_point() +
    labs(title = "GFLOPs across Architectures") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
```

### Visualize energy_per_image vs. F1-score, colored by architecture
```{r energy_vs_f1, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
ggscatter(df_clean,
    x = "energy_per_image", y = "f1_score",
    color = "architecture", facet.by = "training_environment",
    add = "reg.line", conf.int = TRUE,
    palette = "Set1",
) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(title = "energy_per_image vs. F1-score by Architecture and Training Environment")
```

### Efficiency: Megajoule per F1-score
```{r f1_per_joule, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
ggbarplot(df_clean,
    x = "architecture", y = "efficiency",
    fill = "training_environment", position = position_dodge(),
    add = "mean_se", palette = "Set1"
) +
    labs(title = "Efficiency (Megajoule per F1-score) by Architecture and Training Environment")
```

### Visualize trade-offs between F1-score and energy consumption
```{r trade_offs, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
ggplot(df_clean, aes(x = energy, y = f1_score, color = architecture)) +
  geom_point(size = 3) +
  facet_wrap(~ training_environment) +
  labs(
    x = "Total energy consumption (MJ)",
    y = "F1 Score",
    title = "Trade-offs by Environment"
  ) +
  theme_pubr()
```

### Pareto front analysis
```{r pareto_front, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
pf <- h2o.pareto_front(
  df_clean,
  leaderboard_frame = NULL,
  x_metric = "energy",
  y_metric = "f1_score",
  optimum = "top left",
  title = "Pareto Front Plot",
)
plot(pf)
```

### Visualize histograms for F1 score
```{r f1_histograms}
par(mfrow = c(2, 3)) # Set up a 2x3 grid for histograms
hist(df_clean$f1_score, main = "Raw")
hist(log(df_clean$f1_score), main = "Log")
hist(sqrt(df_clean$f1_score), main = "Sqrt")
hist(df_clean$f1_score^2, main = "Square")
hist(df_clean$f1_score^3, main = "Cube")
hist(log(df$f1_score / (1 - df$f1_score)), main = "Logit")
```

## Statistical Analysis
### Model 1: F1-score as a function of architecture and training environment
```{r model1}
df_clean$f1_logit <- log(df_clean$f1_score / (1 - df_clean$f1_score))
model1 <- lmer(f1_score ~ architecture + (1 | training_environment), data = df_clean)
summary(model1)
```
### Model diagnostics for Model 1
```{r model1_diagnostics}
# Check residuals
residuals_model1 <- residuals(model1)
sim_res_model1 <- simulateResiduals(model1)
# Plot residuals
testResiduals(sim_res_model1)
# Check homoscedasticity
homoscedasticity_test <- leveneTest(residuals_model1 ~ architecture, data = df_clean)
homoscedasticity_test
```
```{r check_model1, fig.height=6}
# Check model assumptions
model1_performance <- performance::check_model(model1)
model1_performance
```
```{r}
# Check for outliers
outliers_model1 <- model1_performance$outliers
outliers_model1
# Post hoc tests
post_hoc_model1 <- emmeans(model1, ~ architecture)
post_hoc_results_model1 <- summary(post_hoc_model1)
post_hoc_results_model1
# Plot post hoc results
post_hoc_plot_model1 <- ggplot(post_hoc_results_model1, aes(x = architecture, y = emmean)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, position = position_dodge(0.9)) +
    labs(title = "Post Hoc Results for Model 1", x = "Architecture", y = "Estimated Marginal Mean (F1-score)") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
print(post_hoc_plot_model1)
```

### Model 1 with beta regression
```{r model1_beta}
# Transform F1 to (0,1) if any 0s or 1s
df_clean <- df_clean %>%
    mutate(f1_trans = (f1_score * (n() - 1) + 0.5) / n())

model1_beta <- glmmTMB(f1_trans ~ architecture + (1 | training_environment), 
                      family = beta_family(link = "logit"), 
                      data = df_clean))
summary(model1_beta)
```
```{r model1_beta_diagnostics}
# Check residuals
par(mfrow = c(2, 2))
plot(model1_beta)

sim_beta <- simulateResiduals(model1_beta)
plot(sim_beta)
testUniformity(sim_beta)
```

### Model 2: F1-score per Joule as a function of architecture and training environment
```{r model2}
model2 <- lm(efficiency ~ architecture * training_environment, data = df_clean)
summary(model2)
```
### Model diagnostics for Model 2
```{r model2_diagnostics}
# Check residuals
sim_res_model2 <- simulateResiduals(model2)
# Plot residuals
testResiduals(sim_res_model2)
# Check homoscedasticity
homoscedasticity_test_model2 <- leveneTest(residuals(model2) ~ architecture * training_environment, data = df_clean)
homoscedasticity_test_model2
```
```{r fig.height=10}
# Check model assumptions
model2_performance <- performance::check_model(model2)
model2_performance
```
```{r}
# Check for outliers
outliers_model2 <- model2_performance$outliers
outliers_model2

```{r model2_post_hoc}
# Post hoc tests
post_hoc_model2 <- emmeans(model2, ~ architecture * training_environment)
post_hoc_results_model2 <- summary(post_hoc_model2)
post_hoc_results_model2

pairs(post_hoc_model2, adjust = "tukey") # Post hoc pairwise comparisons

# Plot post hoc results
post_hoc_plot_model2 <- ggplot(post_hoc_results_model2, aes(x = architecture, y = emmean, fill = training_environment)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, position = position_dodge(0.9)) +
    labs(title = "Post Hoc Results for Model 2", x = "Architecture", y = "Estimated Marginal Mean (F1-score per Joule)") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
print(post_hoc_plot_model2)

cld(post_hoc_model2, alpha = 0.05, Letters = letters, adjust = "sidak")
```

## Model Comparison
```{r model_comparison}
# Compare models using ANOVA
# anova(model1, model2, model3)
# # Compare models using AIC
# model_aic <- AIC(model1, model2, model3)
# model_aic <- model_aic %>%
#     mutate(model = c("Model 1", "Model 2", "Model 3")) %>%
#     arrange(AIC)
# # Plot AIC comparison
# ggplot(model_aic, aes(x = model, y = AIC)) +
#     geom_bar(stat = "identity", fill = "steelblue") +
#     labs(title = "AIC Comparison of Models", x = "Model", y = "AIC") +
#     theme_minimal()
# # Compare models using BIC
# model_bic <- BIC(model1, model2, model3)
# model_bic <- model_bic %>%
#     mutate(model = c("Model 1", "Model 2", "Model 3")) %>%
#     arrange(BIC)
# # Plot BIC comparison
# ggplot(model_bic, aes(x = model, y = BIC)) +
#     geom_bar(stat = "identity", fill = "steelblue") +
#     labs(title = "BIC Comparison of Models", x = "Model", y = "BIC") +
#     theme_minimal()
# # Compare models using R-squared
# model_r2 <- data.frame(
#     model = c("Model 1", "Model 2", "Model 3"),
#     r_squared = c(performance::r2(model1), performance::r2(model2), performance::r2(model3))
# )
# # ggplot(model_r2, aes(x = model, y = r_squared)) +
# #     geom_bar(stat = "identity", fill = "steelblue") +
# #     labs(title = "R-squared Comparison of Models", x = "Model", y = "R-squared") +
# #     theme_minimal()
# # Compare models using adjusted R-squared
# model_adj_r2 <- data.frame(
#     model = c("Model 1", "Model 2", "Model 3"),
#     adj_r_squared = c(performance::r2(model1), performance::r2(model2, adjusted = TRUE), performance::r2(model3, adjusted = TRUE))
# )
# # ggplot(model_adj_r2, aes(x = model, y = adj_r_squared)) +
# #     geom_bar(stat = "identity", fill = "steelblue") +
# #     labs(title = "Adjusted R-squared Comparison of Models", x = "Model", y = "Adjusted R-squared") +
# #     theme_minimal()
# # Compare models using RMSE
# model_rmse <- data.frame(
#     model = c("Model 1", "Model 2", "Model 3"),
#     rmse = c(performance::rmse(model1), performance::rmse(model2), performance::rmse(model3))
# )
# ggplot(model_rmse, aes(x = model, y = rmse)) +
#     geom_bar(stat = "identity", fill = "steelblue") +
#     labs(title = "RMSE Comparison of Models", x = "Model", y = "RMSE") +
#     theme_minimal()
# # Compare models using MAE
# model_mae <- data.frame(
#     model = c("Model 1", "Model 2", "Model 3"),
#     mae = c(performance::mae(model1), performance::mae(model2), performance::mae(model3))
# )
# ggplot(model_mae, aes(x = model, y = mae)) +
#     geom_bar(stat = "identity", fill = "steelblue") +
#     labs(title = "MAE Comparison of Models", x = "Model", y = "MAE") +
#     theme_minimal()
# # Compare models using MSE
# model_mse <- data.frame(
#     model = c("Model 1", "Model 2", "Model 3"),
#     mse = c(performance::mse(model1), performance::mse(model2), performance::mse(model3))
# )
# ggplot(model_mse, aes(x = model, y = mse)) +
#     geom_bar(stat = "identity", fill = "steelblue") +
#     labs(title = "MSE Comparison of Models", x = "Model", y = "MSE") +
#     theme_minimal()
```