---
title: "RQ2: Do gains in correctness justify energy-greedy architectures?"
author: "Santiago del Rey"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
    theme: journal
    df_print: paged
knit: internalRmdTools::knit_to_reports
---


```{r setup, include=FALSE}
knitr::opts_knit$set(
    root.dir = "../"
)
library(arrow)
library(lmtest)
library(betareg)
library(gamlss)
library(DHARMa)
library(car)
library(performance) # check assumptions
library(effectsize)
library(emmeans) # post hoc tests
library(multcomp)
library(h2o)
library(gridExtra)
library(ggplot2)
library(ggstar)
library(ggfortify)
library(ggpubr) # for publication-ready plots
library(showtext) # for custom fonts
library(patchwork)
library(tidyverse)
```

```{r setup-fonts, include=FALSE}
# For discrete variables
viridis_scheme <- "turbo"
options(
    ggplot2.discrete.fill = function(...) scale_fill_viridis_d(..., option = viridis_scheme),
    ggplot2.discrete.colour = function(...) scale_colour_viridis_d(..., option = viridis_scheme)
)

# For continuous variables
options(
    ggplot2.continuous.fill = function(...) scale_fill_viridis_c(..., option = viridis_scheme),
    ggplot2.continuous.colour = function(...) scale_colour_viridis_c(..., option = viridis_scheme)
)

font_add("CMU Serif", regular = "cmunrm.ttf") # You need the font installed
showtext_auto()

theme_set(theme_pubr(base_family = "CMU Serif") +
    theme(
        plot.title = element_text(size = 24, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 16),
        strip.text = element_text(size = 18),
        legend.position = "top"
    ))

pub_theme <- theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    strip.text = element_text(size = 10),
)
```

```{r create-figures-directory, include=FALSE}
fig_dir <- "reports/figures/RQ2/"
print(getwd())
if (!dir.exists(fig_dir)) {
    dir.create(fig_dir, recursive = TRUE)
}
```

## Load and preprocess data

We load the data from the parquet file, filter out the `inception_v3` architecture and the `Desktop Normal User` training environment, and normalize the energy consumption by the total number of images seen during training (converting MJ to J). We also prepare the `gpu_usage` variable.

```{r}
df <- read_parquet("data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
    rename(training_environment = `training environment`, gpu_usage = `gpu usage (%)`, f1_score = `f1-score`, energy = `energy (MJ)`) %>%
    filter(architecture != "inception_v3" & training_environment != "Desktop Normal User") %>%
    mutate(
        energy_per_image = (energy * 1000000) / `total seen images`, # Convert MJ to J and normalize by total seen images
        architecture = case_match(
            architecture,
            "mobilenet_v2" ~ "Mobilenet V2",
            "nasnet_mobile" ~ "NASNet Mobile",
            "resnet50" ~ "ResNet50",
            "xception" ~ "Xception",
            "vgg16" ~ "VGG16"
        ),
        architecture = factor(architecture, levels = c("Mobilenet V2", "NASNet Mobile", "ResNet50", "Xception", "VGG16")),
        GFLOPs = factor(GFLOPs),
        training_environment = factor(training_environment, levels = c("Desktop ML Engineer", "Server")),
        efficiency = (energy / max(energy, na.rm = TRUE)) / (f1_score / max(f1_score, na.rm = TRUE)), # Megajoule per F1-score
    ) %>%
    dplyr::select(energy, energy_per_image, f1_score, efficiency, GFLOPs, architecture, training_environment, `training duration (h)`)
```

### Outlier filtering functions

```{r}
# Define a function to filter outliers of a given variable on an input data frame
filter_outliers <- function(df, variable) {
    df %>%
        group_by("GFLOPs", "training_environment") %>%
        mutate(
            Q1 = quantile(!!sym(variable), 0.25),
            Q3 = quantile(!!sym(variable), 0.75),
            IQR = Q3 - Q1,
            is_outlier = !!sym(variable) < (Q1 - 1.5 * IQR) | !!sym(variable) > (Q3 + 1.5 * IQR)
        ) %>%
        ungroup() %>%
        filter(!is_outlier) %>%
        dplyr::select(-Q1, -Q3, -IQR, -is_outlier)
}
```

## Data overview
```{r data_overview}
# Filter F1-score outliers
df_clean <- df %>%
    group_by(GFLOPs) %>%
    mutate(
        Q1 = quantile(f1_score, 0.25),
        Q3 = quantile(f1_score, 0.75),
        IQR = Q3 - Q1,
        is_outlier = f1_score < (Q1 - 1.5 * IQR) | f1_score > (Q3 + 1.5 * IQR)
    ) %>%
    ungroup() %>%
    dplyr::select(-Q1, -Q3, -IQR)

df_clean %>%
    filter(is_outlier)

df_clean <- df_clean %>%
    filter(!is_outlier) %>%
    dplyr::select(-is_outlier)

nrows <- nrow(df_clean)
df_clean <- filter_outliers(df_clean, "energy_per_image")
print(paste("Number of outliers removed from energy_per_image:", nrows - nrow(df_clean)))

df_clean %>%
    filter(f1_score < 0.5)


# Display the first few rows of the data frame
head(df_clean)
# Display summary statistics of the data frame
summary(df_clean)

df_clean %>%
    group_by(architecture) %>%
    summarise(
        min_f1_score = min(f1_score, na.rm = TRUE),
        max_f1_score = max(f1_score, na.rm = TRUE),
        mean_f1_score = mean(f1_score, na.rm = TRUE),
        sd_f1_score = sd(f1_score, na.rm = TRUE),
        median_f1_score = median(f1_score, na.rm = TRUE),
        min_energy_per_image = min(energy_per_image, na.rm = TRUE),
        max_energy_per_image = max(energy_per_image, na.rm = TRUE),
        mean_energy_per_image = mean(energy_per_image, na.rm = TRUE),
        sd_energy_per_image = sd(energy_per_image, na.rm = TRUE),
        median_energy_per_image = median(energy_per_image, na.rm = TRUE),
        min_efficiency = min(efficiency, na.rm = TRUE),
        max_efficiency = max(efficiency, na.rm = TRUE),
        mean_efficiency = mean(efficiency, na.rm = TRUE),
        sd_efficiency = sd(efficiency, na.rm = TRUE),
        median_efficiency = median(efficiency, na.rm = TRUE)
    ) %>%
    ungroup() %>%
    arrange(architecture)
```

## Exploratory visualization
### Visualize the GFLOPs of each architecture
```{r gflops_by_arch, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
ggplot(df_clean %>% distinct(architecture, .keep_all = TRUE), aes(x = architecture, y = GFLOPs)) +
    geom_col() +
    labs(x = "Architecture", y = "GFLOPs")
```

### F1-score and energy consumption by GFLOPs and training environment
```{r f1_by_arch_env, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
f1_boxplot <- ggboxplot(df_clean,
    x = "architecture", y = "f1_score",
) +
    labs(x = "Architecture", y = "F1-score")

plot(f1_boxplot)

ggsave(
    plot = f1_boxplot + pub_theme,
    filename = file.path(fig_dir, "f1_by_arch.pdf"),
    device = cairo_pdf,
    units = "cm",
    width = 13.71, height = 8, dpi = 300
)

ggboxplot(df_clean,
    x = "architecture", y = "energy_per_image",
    fill = "training_environment",
) +
    labs(x = "Architecture", y = "Energy consumption (J/image)", fill = "Training Environment")
```

### Visualize energy_per_image vs. F1-score, colored by GFLOPs
```{r energy_vs_f1, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
ggscatter(df_clean,
    x = "energy_per_image", y = "f1_score",
    color = "GFLOPs", facet.by = "training_environment",
    add = "reg.line", conf.int = TRUE,
    palette = "Set1",
) +
    geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
    coord_cartesian(ylim = c(0, 1)) +
    labs(title = "Joules/Image vs. F1-score by GFLOPs and Training Environment")
```

### Efficiency: Megajoule per F1-score
```{r f1_per_joule, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
ggbarplot(df_clean,
    x = "GFLOPs", y = "efficiency",
    fill = "training_environment", position = position_dodge(),
    add = "mean_se", palette = "Set1"
) +
    labs(title = "Efficiency (Megajoule per F1-score) by GFLOPs and Training Environment")
```

### Visualize trade-offs between F1-score and energy consumption
```{r trade_offs, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
trade_off_plot <- ggplot(df_clean, aes(x = energy, y = f1_score, color = GFLOPs)) +
    geom_point(size = 1) +
    facet_wrap(~training_environment) +
    labs(
        x = "Total energy consumption (MJ)",
        y = "F1 Score",
    ) +
    scale_color_discrete(labels = function(x) round(as.numeric(x), 2))

plot(trade_off_plot)

ggsave(
    plot = trade_off_plot + pub_theme,
    filename = file.path(fig_dir, "trade_offs_f1_vs_energy.pdf"),
    device = cairo_pdf,
    units = "cm",
    width = 13.71, height = 8, dpi = 300
)
```

### Pareto front analysis
```{r pareto_front, fig.align="center", echo=FALSE, fig.width=10, fig.height=6}
pf <- h2o.pareto_front(
    df_clean,
    leaderboard_frame = NULL,
    x_metric = "energy",
    y_metric = "f1_score",
    optimum = "top left",
    title = "Pareto Front Plot",
)

# Get list of Pareto optimal points
pf_points_df <- pf@pareto_front

pf_points_df

pareto_plot <- ggplot(df_clean, aes(x = energy, y = f1_score, color = architecture, shape = training_environment)) +
    geom_point(size = 1, alpha = 0.7) +
    geom_line(data = pf_points_df, aes(x = energy, y = f1_score, group = 1), linewidth = 1.1, alpha = 0.8, color = "black") +
    geom_star(data = pf_points_df, aes(x = energy, y = f1_score, fill = architecture, group = 1), size = 1.5, alpha = 1, show.legend = FALSE) +
    labs(
        x = "Total energy consumption (MJ)",
        y = "F1 Score",
    ) +
    guides(color = guide_legend(title = "Architecture", nrow = 3), shape = guide_legend(title = "Training Environment"), fill = FALSE) +
    theme(
        legend.position = "inside",
        legend.justification = c("right", "bottom"),
        legend.box.just = "left",
        legend.margin = margin(0, 0, 0, 0),
        legend.background = element_rect(fill = "transparent"),
    )

plot(pareto_plot)

ggsave(
    plot = pareto_plot + pub_theme,
    filename = file.path(fig_dir, "pareto_front_f1_vs_energy.pdf"),
    device = cairo_pdf,
    units = "cm",
    width = 13.71, height = 8, dpi = 300
)
```

### Visualize histograms for F1 score
```{r f1_histograms}
par(mfrow = c(2, 3)) # Set up a 2x3 grid for histograms
hist(df_clean$f1_score, main = "Raw")
hist(log(df_clean$f1_score), main = "Log")
hist(sqrt(df_clean$f1_score), main = "Sqrt")
hist(df_clean$f1_score^2, main = "Square")
hist(df_clean$f1_score^3, main = "Cube")
hist(log(df$f1_score / (1 - df$f1_score)), main = "Logit")
```

## Statistical Analysis

### Correlation between F1-score and energy consumption
```{r correlation_f1_energy}
cor.test(df_clean$f1_score, df_clean$energy_per_image, method = "spearman")
```

### Model 1: F1-score as a function of GFLOPs
```{r model1}
df_clean$f1_logit <- log(df_clean$f1_score / (1 - df_clean$f1_score))
model1 <- lm(f1_logit ~ GFLOPs, data = df_clean)
summary(model1)
```
### Model diagnostics for Model 1
```{r model1_diagnostics}
# Check residuals
residuals_model1 <- stats::residuals(model1)
sim_res_model1 <- simulateResiduals(model1)
# Plot residuals
testResiduals(sim_res_model1)
```
```{r check_model1, fig.height=6}
# Check model assumptions
model1_performance <- performance::check_model(model1)
model1_performance
```
```{r model1_outliers}
# Check for outliers
outliers_model1 <- model1_performance$outliers
outliers_model1
```
```{r model1_post_hoc}
# Post hoc tests
post_hoc_model1 <- emmeans(model1, ~GFLOPs)
post_hoc_results_model1 <- summary(post_hoc_model1)
post_hoc_results_model1
# Plot post hoc results
post_hoc_plot_model1 <- ggplot(post_hoc_results_model1, aes(x = GFLOPs, y = emmean)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, position = position_dodge(0.9)) +
    labs(title = "Post Hoc Results for Model 1", x = "GFLOPs", y = "Estimated Marginal Mean (F1-score)") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
print(post_hoc_plot_model1)

pairs(post_hoc_model1, adjust = "tukey") # Post hoc pairwise comparisons
```

### Model 1 with beta regression
```{r model1_beta}
# Transform F1 to (0,1) if any 0s or 1s
df_clean <- df_clean %>%
    group_by(GFLOPs) %>%
    mutate(f1_trans = (f1_score * (n() - 1) + 0.5) / n()) %>%
    ungroup()

# Original model
model1_beta <- betareg(f1_trans ~ GFLOPs, data = df_clean)
summary(model1_beta)
```

```{r model1_beta_diagnostics}
# Check residuals
par(mfrow = c(2, 2))
plot(model1_beta)

resid_quant <- residuals(model1_beta, type = "quantile") # Quantile residuals
leverage <- hatvalues(model1_beta) # Influence measures

# Cook's distance analogue (optional, not always directly provided)
# If not available: approximate influence = leverage * (residual^2)
influence_metric <- leverage * (resid_quant^2)

# Combine into a data frame
diagnostics_df <- df_clean %>%
    mutate(
        residuals = resid_quant,
        leverage = leverage,
        influence = influence_metric,
        obs_id = row_number()
    )

# Heuristics
high_leverage <- diagnostics_df$leverage > (2 * mean(diagnostics_df$leverage))
high_residuals <- abs(diagnostics_df$residuals) > 2
high_influence <- diagnostics_df$influence > (2 * mean(diagnostics_df$influence))

# Label flags
diagnostics_df <- diagnostics_df %>%
    mutate(flag = case_when(
        high_leverage & high_residuals ~ "High leverage & residual",
        high_influence ~ "High influence",
        TRUE ~ "Normal"
    ))


ggplot(diagnostics_df, aes(x = leverage, y = residuals, color = flag)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed") +
    labs(title = "Residuals vs Leverage", x = "Leverage", y = "Quantile Residuals")

ggplot(diagnostics_df, aes(x = obs_id, y = influence)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    geom_hline(yintercept = 2 * mean(diagnostics_df$influence), color = "red", linetype = "dashed") +
    labs(title = "Influence Metric (Cook's Distance Approx.)", x = "Observation", y = "Influence")

diagnostics_df %>%
    arrange(desc(influence)) %>%
    dplyr::select(obs_id, GFLOPs, training_environment, residuals, leverage, influence, flag) %>%
    head(10)

df_clean %>%
    filter(row_number() %in% c(198, 219, 237, 202, 204, 230, 215, 214, 192, 212))

ggplot(filter(diagnostics_df, GFLOPs == "ResNet50"), aes(x = obs_id, y = residuals)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed") +
    labs(title = "Residuals for ResNet50")
```

```{r model1_post_hoc_beta}
# Post hoc tests for beta regression model
post_hoc_model1_beta <- emmeans(model1_beta, ~GFLOPs, type = "response")
post_hoc_results_model1_beta <- summary(post_hoc_model1_beta)
post_hoc_results_model1_beta
# Plot post hoc results for beta regression model
post_hoc_plot_model1_beta <- ggplot(post_hoc_results_model1_beta, aes(x = GFLOPs, y = emmean)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    # geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, position = position_dodge(0.9)) +
    labs(title = "Post Hoc Results for Model 1 (Beta Regression)", x = "GFLOPs", y = "Estimated Marginal Mean (F1-score)") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
print(post_hoc_plot_model1_beta)

pairs(post_hoc_model1_beta, adjust = "tukey") # Post hoc pairwise comparisons for beta regression model
```


### Model 1 using a GAMLSS
```{r model1_gamlss}
# Full model
full_model <- gamlss(
    f1_score ~ GFLOPs,
    sigma.formula = ~GFLOPs,
    family = BE,
    control = gamlss.control(n.cyc = 100),
    data = df_clean
)

summary(full_model)

plot(full_model)

null_model <- gamlss(f1_score ~ 1, sigma.formula = ~1, family = BE, data = df_clean)

drop1(full_model)

model_env_only <- gamlss(f1_score ~ training_environment, sigma.formula = ~1, family = BE, data = df_clean)
drop1(model_env_only)
```

```{r model1_gamlss_effects}
# Get unique combinations of factors
# Extract predicted means and SE on response scale (mu)
emm <- emmeans(full_model, ~GFLOPs,
    type = "response",
    at = list(sigma = mean(predict(full_model, what = "sigma", type = "response")))
)

# Convert emmeans to dataframe for ggplot
emm_df <- as.data.frame(emm)
emm_df

ggplot(emm_df, aes(x = GFLOPs, y = response)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.2) +
    ylab("Estimated mean F1 score") +
    xlab("GFLOPs") +
    ggtitle("Effect of Model GFLOPs on F1 Score") +
    theme_minimal()
```

```{r model1_gamlss_post_hoc}
pairs_emm <- pairs(emm, adjust = "tukey")
pairs_df <- as.data.frame(pairs_emm)
pairs_df

ggplot(pairs_df, aes(x = contrast, y = odds.ratio)) +
    geom_point() +
    geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
    coord_flip() +
    ylab("Odds Ratio (F1 score)") +
    xlab("GFLOPs contrasts") +
    ggtitle("Pairwise Comparisons of GFLOPs (Tukey adjusted)") +
    theme_minimal()
```

### Model 2: F1-score per Joule as a function of GFLOPs and training environment
```{r model2}
model2 <- lm(efficiency ~ GFLOPs * training_environment, data = df_clean)
summary(model2)
```
#### Model diagnostics for Model 2
```{r model2_diagnostics}
# Check residuals
sim_res_model2 <- simulateResiduals(model2)
# Plot residuals
testResiduals(sim_res_model2)
```
```{r check-model2, fig.width=8, fig.height=10}
check_model(model2)
```

#### Sifgnificance of interaction effects
```{r model2_interaction}
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model2, type = 3)
```

#### Post hoc tests for Model 2
```{r model2_post_hoc}
# Post hoc tests
post_hoc_model2 <- emmeans(model2, ~ GFLOPs * training_environment)
post_hoc_results_model2 <- summary(post_hoc_model2)
post_hoc_results_model2

pairs(post_hoc_model2, adjust = "tukey") # Post hoc pairwise comparisons

# Plot post hoc results
post_hoc_plot_model2 <- ggplot(post_hoc_results_model2, aes(x = GFLOPs, y = emmean, fill = training_environment)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, position = position_dodge(0.9)) +
    labs(title = "Post Hoc Results for Model 2", x = "GFLOPs", y = "Estimated Marginal Mean (F1-score per Joule)") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
print(post_hoc_plot_model2)

cld(post_hoc_model2, alpha = 0.05, Letters = letters, adjust = "sidak")
```
