---
title: "RQ3: Is there a relationship between GPU usage and energy consumption when training DL models?"
author: "Santiago del Rey"
output: html_notebook
---
## Load libraries
```{r setup, include=FALSE}
library(arrow)
library(dplyr)
library(lme4)
library(lmerTest) # for p-values in mixed models
library(performance)
library(DHARMa)
library(car) # for Anova function
library(ggplot2)
library(ggpubr) # for publication-ready plots
```

## Load data
```{r load-data}
df <- read_parquet("../data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
    rename(
        training_environment = `training environment`,
        gpu_usage = `gpu usage (%)`,
        energy = `energy (MJ)`,
        power_draw = `average gpu power (W)`,
        tdp = `max power limit (W)`,
    ) %>%
    filter(architecture != "inception_v3" & training_environment != "Desktop Normal User") %>%
    mutate(
        energy_per_image = (energy * 1000000) / `total seen images`, # Convert MJ to KJ and normalize by total seen images
        gpu_usage = gpu_usage / 100, # Convert percentage to proportion
        architecture = case_match(
            architecture,
            "mobilenet_v2" ~ "Mobilenet V2",
            "nasnet_mobile" ~ "NASNet Mobile",
            "resnet50" ~ "ResNet50",
            "xception" ~ "Xception",
            "vgg16" ~ "VGG16"
        ),
        architecture = factor(architecture, levels = c("Mobilenet V2", "NASNet Mobile", "ResNet50", "Xception", "VGG16")),
        training_environment = factor(training_environment, levels = c("Desktop ML Engineer", "Server")),
    ) %>%
    select(training_environment, architecture, gpu_usage, energy_per_image, power_draw, tdp)

filter_outliers <- function(df, variable) {
    df %>%
        group_by(architecture, training_environment) %>%
        mutate(
            Q1 = quantile(!!sym(variable), 0.25),
            Q3 = quantile(!!sym(variable), 0.75),
            IQR = Q3 - Q1,
            is_outlier = !!sym(variable) < (Q1 - 1.5 * IQR) | !!sym(variable) > (Q3 + 1.5 * IQR)
        ) %>%
        ungroup() %>%
        filter(!is_outlier) %>%
        dplyr::select(-Q1, -Q3, -IQR, -is_outlier)
}

df_clean_energy <- filter_outliers(df, "gpu_usage") %>%
    filter_outliers("energy_per_image")

df_clean_power <- filter_outliers(df, "gpu_usage") %>%
    filter_outliers("power_draw")
```

## Plot GPU usage vs energy consumption
```{r gpu-usage-vs-energy-consumption}
ggscatter(df_clean_energy,
    x = "gpu_usage", y = "energy_per_image",
    color = "training_environment",
    size = 3, alpha = 0.7,
    xlab = "GPU Usage (%)", ylab = "Normalized Energy Consumption (J/image)",
    title = "GPU Usage vs Energy Consumption in DL Training"
) +
    geom_smooth(method = "lm", se = FALSE, aes(color = training_environment)) +
    facet_wrap(~training_environment) +
    theme(legend.position = "top")
```

## Plot GPU usage vs average power draw
```{r gpu-usage-vs-power-draw}
ggscatter(df_clean_power,
    x = "gpu_usage", y = "power_draw",
    color = "training_environment",
    size = 3, alpha = 0.7,
    xlab = "GPU Usage (%)", ylab = "Average Power Draw (W)",
    title = "GPU Usage vs Average Power Draw in DL Training"
) +
    # Add a line for the TDP * GPU usage
    geom_line(aes(y = gpu_usage * tdp), linetype = "dashed", color = "black") +
    geom_text(aes(label = paste0("TDP: ", tdp, " W")), x = min(df_clean_power$gpu_usage) + 10, y = max(df_clean_power$power_draw) * 0.9, color = "red") +
    geom_smooth(method = "lm", se = FALSE, aes(color = training_environment)) +
    facet_wrap(~training_environment) +
    theme(legend.position = "top")
```

## Statistical analysis
### Energy consumption analysis
#### Global linear trend
```{r global-linear-trend}
model_linear <- lm(energy_per_image ~ gpu_usage, data = df_clean_energy)
summary(model_linear)
anova(model_linear)
```

##### Model diagnostics
```{r check-residuals-linear}
# Check residuals
par(mfrow = c(2, 2))
plot(model_linear)
```
```{r check-dharma-linear}
# Check residuals with DHARMa
sim_linear <- simulateResiduals(model_linear)
testResiduals(sim_linear)
```
```{r check-model-linear, fig.width=8, fig.height=10}
check_model(model_linear)
```

#### Interaction effects
```{r statistical-analysis}
# Perform linear regression
model <- lm(energy_per_image ~ gpu_usage * training_environment * architecture, data = df_clean_energy)
summary(model)
# ANOVA to check for significance
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model, type = 3)
```

##### Model diagnostics
```{r check-residuals}
# Check residuals
par(mfrow = c(2, 2))
plot(model)
```

```{r check-dharma}
# Check residuals with DHARMa
sim <- simulateResiduals(model)
testResiduals(sim)
```

#### Mixed effects model
```{r mixed-effects-model}
model_mixed <- lmer(energy_per_image ~ gpu_usage + (1 | architecture) + (1 | training_environment), data = df_clean_energy)
summary(model_mixed)
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_mixed, type = 3)
```

##### Model diagnostics
```{r check-residuals-mixed}
# Check residuals
par(mfrow = c(2, 2))
plot(model_mixed)
```
```{r check-dharma-mixed}
# Check residuals with DHARMa
sim_mixed <- simulateResiduals(model_mixed)
testResiduals(sim_mixed)
```

### Power draw analysis
#### Global linear trend
```{r global-linear-trend-power-draw}
model_linear_power_draw <- lm(power_draw ~ gpu_usage, data = df_clean_power)
summary(model_linear_power_draw)
anova(model_linear_power_draw)
```
##### Model diagnostics for power draw
```{r check-residuals-linear-power-draw}
# Check residuals
par(mfrow = c(2, 2))
plot(model_linear_power_draw)
```
```{r check-dharma-linear-power-draw}
# Check residuals with DHARMa
sim_linear_power_draw <- simulateResiduals(model_linear_power_draw)
testResiduals(sim_linear_power_draw)
```

#### Interaction effects for power draw
```{r model-power-draw}
model_power_draw <- lm(power_draw ~ gpu_usage * training_environment * architecture, data = df_clean_power)
summary(model_power_draw)
```

```{r anova-power-draw}
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_power_draw, type = 3)
```

##### Model diagnostics for power draw
```{r check-residuals-power-draw}
# Check residuals
par(mfrow = c(2, 2))
plot(model_power_draw)
```
```{r check-dharma-power-draw}
# Check residuals with DHARMa
sim_power_draw <- simulateResiduals(model_power_draw)
testResiduals(sim_power_draw)
```
```{r check-model-power-draw, fig.width=8, fig.height=10}
check_model(model_power_draw)
```

#### Mixed effects model for power draw
```{r mixed-effects-model-power-draw}
model_mixed_power_draw <- lmer(power_draw ~ gpu_usage * training_environment * architecture + (1 | tdp), data = df_clean_power)
summary(model_mixed_power_draw)
```

```{r anova-mixed-power-draw}
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_mixed_power_draw, type = 3)
```
##### Model diagnostics for mixed effects model power draw
```{r check-residuals-mixed-power-draw}
# Check residuals
par(mfrow = c(2, 2))
plot(model_mixed_power_draw)
```
```{r check-dharma-mixed-power-draw}
# Check residuals with DHARMa
sim_mixed_power_draw <- simulateResiduals(model_mixed_power_draw)
testResiduals(sim_mixed_power_draw)
```
```{r check-model-mixed-power-draw, fig.width=8, fig.height=10}
check_model(model_mixed_power_draw)
```
