---
title: "RQ3: Is there a relationship between GPU usage and energy consumption when training DL models?"
author: "Santiago del Rey"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: journal
    df_print: paged
knit: internalRmdTools::knit_to_reports
---
```{r setup, include=FALSE}
knitr::opts_knit$set(
    root.dir = "../"
)
library(arrow)
library(lme4)
library(lmerTest) # for p-values in mixed models
library(performance)
library(DHARMa)
library(car) # for Anova function
library(emmeans) # for post-hoc tests
library(effectsize)
library(multcomp)
library(ggplot2)
library(ggpubr) # for publication-ready plots
library(showtext)
library(patchwork)
library(tidyverse)
```

```{r setup-fonts, include=FALSE}
# For discrete variables
viridis_scheme <- "turbo"
options(
    ggplot2.discrete.fill = function(...) scale_fill_viridis_d(..., option = viridis_scheme),
    ggplot2.discrete.colour = function(...) scale_colour_viridis_d(..., option = viridis_scheme)
)

# For continuous variables
options(
    ggplot2.continuous.fill = function(...) scale_fill_viridis_c(..., option = viridis_scheme),
    ggplot2.continuous.colour = function(...) scale_colour_viridis_c(..., option = viridis_scheme)
)

font_add("CMU Serif", regular = "cmunrm.ttf") # You need the font installed
showtext_auto()

theme_set(theme_pubr(base_family = "CMU Serif") +
    theme(
        plot.title = element_text(size = 24, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 18),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 16),
        strip.text = element_text(size = 18),
        legend.position = "top"
    ))

pub_theme <- theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    strip.text = element_text(size = 10),
)
```

```{r create-figures-directory, include=FALSE}
fig_dir <- "reports/figures/RQ3/"
print(getwd())
if (!dir.exists(fig_dir)) {
    dir.create(fig_dir, recursive = TRUE)
}
```

## Load data
```{r load-data}
df <- read_parquet("data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
    rename(
        training_environment = `training environment`,
        gpu_usage = `gpu usage (%)`,
        energy = `energy (MJ)`,
        power_draw = `average gpu power (W)`,
        tdp = `max power limit (W)`,
    ) %>%
    filter(architecture != "inception_v3" & training_environment != "Desktop Normal User") %>%
    mutate(
        energy_per_image = (energy * 1000000) / `total seen images`, # Convert MJ to KJ and normalize by total seen images
        gpu_usage = gpu_usage / 100, # Convert percentage to proportion
        architecture = case_match(
            architecture,
            "mobilenet_v2" ~ "Mobilenet V2",
            "nasnet_mobile" ~ "NASNet Mobile",
            "resnet50" ~ "ResNet50",
            "xception" ~ "Xception",
            "vgg16" ~ "VGG16"
        ),
        architecture = factor(architecture, levels = c("Mobilenet V2", "NASNet Mobile", "ResNet50", "Xception", "VGG16")),
        training_environment = factor(training_environment, levels = c("Desktop ML Engineer", "Server")),
    ) %>%
    select(training_environment, architecture, gpu_usage, energy_per_image, power_draw, tdp)

filter_outliers <- function(df, variable) {
    df %>%
        group_by(architecture, training_environment) %>%
        mutate(
            Q1 = quantile(!!sym(variable), 0.25),
            Q3 = quantile(!!sym(variable), 0.75),
            IQR = Q3 - Q1,
            is_outlier = !!sym(variable) < (Q1 - 1.5 * IQR) | !!sym(variable) > (Q3 + 1.5 * IQR)
        ) %>%
        ungroup() %>%
        filter(!is_outlier) %>%
        dplyr::select(-Q1, -Q3, -IQR, -is_outlier)
}

df_clean_energy <- filter_outliers(df, "gpu_usage") %>%
    filter_outliers("energy_per_image")

df_clean_power <- filter_outliers(df, "gpu_usage") %>%
    filter_outliers("power_draw")
```

## Plot GPU usage vs energy consumption
```{r gpu-usage-vs-energy-consumption}
gpu_vs_energy <- ggplot(df_clean_energy,
    aes(x = gpu_usage, y = energy_per_image, color = training_environment),
    size = 3, alpha = 0.7
) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
        x = "GPU Usage (%)",
        y = "Normalized Energy (J/image)",
    ) +
    guides(
        color = guide_legend(title = "Training environment")
    )

gpu_vs_energy

ggsave(
    plot = gpu_vs_energy + pub_theme,
    filename = file.path(fig_dir, "gpu_usage_vs_energy_consumption.pdf"),
    device = cairo_pdf,
    units = "cm",
    width = 13.71, height = 8, dpi = 300
)
```

## Plot GPU usage vs average power draw
```{r gpu-usage-vs-power-draw}
server_tdp <- df_clean_power %>%
    filter(training_environment == "Server") %>%
    summarise(tdp = mean(tdp, na.rm = TRUE)) %>%
    pull(tdp)

desktop_tdp <- df_clean_power %>%
    filter(training_environment == "Desktop ML Engineer") %>%
    summarise(tdp = mean(tdp, na.rm = TRUE)) %>%
    pull(tdp)

gpu_vs_power <- ggplot(
    df_clean_power,
    aes(x = gpu_usage, y = power_draw, color = training_environment),
    size = 3, alpha = 0.7
) +
    geom_point() +
    geom_line(aes(y = gpu_usage * tdp, linetype = "dashed")) +
    geom_smooth(method = "lm", se = FALSE, aes(linetype = "solid")) +
    # Add an element indicating that dashed lines represent TDP * GPU usage
    scale_linetype_manual(
        name = element_blank(),
        values = c("dashed", "solid"),
        labels = c("TDP x GPU usage", "Fit")
    ) +
    labs(
        x = "GPU Usage (%)",
        y = "Average Power Draw (W)",
    ) +
    guides(
        color = guide_legend(title = element_blank(), order = 1),
        linetype = guide_legend(order = 2)
    ) +
    theme(legend.direction = "vertical")


gpu_vs_power

ggsave(
    plot = gpu_vs_power + pub_theme,
    filename = file.path(fig_dir, "gpu_usage_vs_power_draw.pdf"),
    device = cairo_pdf,
    units = "cm",
    width = 13.71, height = 8, dpi = 300
)
```

```{r gpu-usage-vs-enegy-and-power}
gpu_vs_energy <- gpu_vs_energy +
    guides(color = "none")

gpu_usage_vs_power_and_energy <- (gpu_vs_energy | gpu_vs_power) +
    plot_layout(axes = "collect", guides = "collect")

ggsave(
    plot = gpu_usage_vs_power_and_energy & pub_theme,
    filename = file.path(fig_dir, "gpu_usage_vs_power_draw_and_energy.pdf"),
    device = cairo_pdf,
    units = "cm",
    width = 13.71, height = 9, dpi = 300
)
```

## Statistical analysis
### Energy consumption analysis
#### Global linear trend
```{r global-linear-trend}
model_global <- lm(energy_per_image ~ gpu_usage, data = df_clean_energy)
summary(model_global)
anova(model_global)
```

##### Model diagnostics
```{r check-residuals-linear}
# Check residuals
par(mfrow = c(2, 2))
plot(model_global)
```
```{r check-dharma-linear}
# Check residuals with DHARMa
sim_linear <- simulateResiduals(model_global)
testResiduals(sim_linear)
```
```{r check-model-linear, fig.width=8, fig.height=10}
check_model(model_global)
```

#### Interaction effects
```{r statistical-analysis}
# Perform linear regression
model_linear <- lm(energy_per_image ~ gpu_usage * training_environment * architecture, data = df_clean_energy)
summary(model_linear)
# ANOVA to check for significance
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_linear, type = 3)

eta_squared(model_linear, partial = FALSE)      # Eta² (total variance explained)
eta_squared(model_linear, partial = TRUE)       # Partial Eta² (variance explained after controlling for others)
```

##### Model diagnostics
```{r check-linear}
check_model(model_linear)
```

```{r check-dharma}
# Check residuals with DHARMa
sim <- simulateResiduals(model_linear)
testResiduals(sim)
```

#### Mixed effects model
```{r mixed-effects-model}
model_mixed <- lmer(energy_per_image ~ gpu_usage + (1 | architecture) + (1 | training_environment), data = df_clean_energy)
summary(model_mixed)
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_mixed, type = 3)
```

##### Model diagnostics
```{r check-residuals-mixed}
# Check residuals
par(mfrow = c(2, 2))
plot(model_mixed)
```
```{r check-dharma-mixed}
# Check residuals with DHARMa
sim_mixed <- simulateResiduals(model_mixed)
testResiduals(sim_mixed)
```

#### Compare models
```{r compare-models}
compare_performance(model_global, model_linear, model_mixed, rank = TRUE)
```

#### Post-hoc tests
```{r post-hoc-tests}
# Estimate marginal means with GPU usage at 50% (0.5)
emms <- emmeans(model_linear, ~ gpu_usage * training_environment * architecture, at = list(gpu_usage = 0.9))
emms
pairwise_com <- pairs(emms)
pairwise_com

# Select significant comparisons
as.data.frame(pairwise_com) %>%
    filter(p.value < 0.05)

compact_letter_display <- cld(emms, Letters = letters, adjust = "sidak")

print(compact_letter_display)

ggplot(compact_letter_display, aes(
    x = interaction(gpu_usage, training_environment, architecture),
    y = emmean
)) +
    geom_col(fill = "skyblue") +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
    geom_text(aes(label = .group), vjust = -0.5, color = "red") +
    labs(x = "Condition", y = "Estimated Marginal Mean (GPU Usage)") +
    theme_pubr() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Power draw analysis
#### Global linear trend
```{r global-linear-trend-power-draw}
model_global_power_draw <- lm(power_draw ~ gpu_usage, data = df_clean_power)
summary(model_global_power_draw)
anova(model_global_power_draw)
```
##### Model diagnostics for power draw
```{r check-residuals-linear-power-draw}
# Check residuals
par(mfrow = c(2, 2))
plot(model_global_power_draw)
```
```{r check-dharma-linear-power-draw}
# Check residuals with DHARMa
sim_linear_power_draw <- simulateResiduals(model_global_power_draw)
testResiduals(sim_linear_power_draw)
```

#### Interaction effects for power draw
```{r model-power-draw}
model_linear_power_draw <- lm(power_draw ~ gpu_usage * training_environment * architecture, data = df_clean_power)
summary(model_linear_power_draw)
```

```{r anova-power-draw}
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_linear_power_draw, type = 3)

eta_squared(model_linear, partial = FALSE)      # Eta² (total variance explained)
eta_squared(model_linear, partial = TRUE)       # Partial Eta² (variance explained after controlling for others)
```

##### Model diagnostics for power draw
```{r check-residuals-power-draw}
# Check residuals
par(mfrow = c(2, 2))
plot(model_linear_power_draw)
```
```{r check-dharma-power-draw}
# Check residuals with DHARMa
sim_power_draw <- simulateResiduals(model_linear_power_draw)
testResiduals(sim_power_draw)
```
```{r check-model-power-draw, fig.width=8, fig.height=10}
check_model(model_linear_power_draw)
```

#### Mixed effects model for power draw
```{r mixed-effects-model-power-draw}
model_mixed_power_draw <- lmer(power_draw ~ gpu_usage * training_environment * architecture + (1 | tdp), data = df_clean_power)
summary(model_mixed_power_draw)
```

```{r anova-mixed-power-draw}
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_mixed_power_draw, type = 3)
```
##### Model diagnostics for mixed effects model power draw
```{r check-residuals-mixed-power-draw}
# Check residuals
par(mfrow = c(2, 2))
plot(model_mixed_power_draw)
```
```{r check-dharma-mixed-power-draw}
# Check residuals with DHARMa
sim_mixed_power_draw <- simulateResiduals(model_mixed_power_draw)
testResiduals(sim_mixed_power_draw)
```
```{r check-model-mixed-power-draw, fig.width=8, fig.height=10}
check_model(model_mixed_power_draw)
```

#### Compare models for power draw
```{r compare-models-power-draw}
compare_performance(model_global_power_draw, model_linear_power_draw, model_mixed_power_draw, rank = TRUE)
```

#### Post-hoc tests for power draw
```{r post-hoc-tests-power-draw}
emms_power_draw <- emmeans(model_linear_power_draw, ~ gpu_usage * training_environment * architecture)
emms_power_draw

pairs(emms_power_draw, adjust = "tukey")

plot(emms_power_draw, comparisons = TRUE)
```
