---
title: "RQ3"
output: html_notebook
author: "Santiago del Rey"
---

## Load libraries
```{r include=FALSE}
library(arrow)
library(dplyr)
library(ggplot2)
library(ggpubr) # for publication-ready plots
library(lme4)
library(nlme)
library(lmerTest) # p-values for lmer
library(performance) # check assumptions
library(emmeans) # post hoc tests
library(car)
library(rstatix)
library(effectsize)
library(minpack.lm)
library(boot)
```

## Load data
```{r}
df <- read_parquet("../data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
  filter(architecture == "inception_v3" & `training environment` == "Server") %>%
  mutate(
    energy = (`energy (MJ)` * 1000000) / `total seen images`, # Convert MJ to J and normalize by total seen images
    batch_size = factor(`batch size`),
    input_size = factor(`image size`),
    dataset = factor(`dataset`),
    gpu_usage = `gpu usage (%)`,
  ) %>%
  group_by(batch_size, input_size, dataset) %>%
  mutate(
    subject_id = factor(row_number()),
  )
```

## Data summary
```{r}
df %>%
  group_by(batch_size, input_size, dataset) %>%
  get_summary_stats(energy, type = "mean_sd")
```

## Visualize data
```{r}
bxp <- ggboxplot(
  df,
  x = "input_size", y = "energy",
  color = "batch_size", palette = "jco"
)
bxp
```

## Create a QQ plot for each cell of the design
```{r}
ggqqplot(df, "energy", ggtheme = theme_bw()) +
  facet_grid(batch_size ~ input_size, labeller = "label_both")
```

## Create a linear mixed model
Create a linear mixed model to analyze the energy consumption based on batch size and input size, with random effects for subject and dataset.
```{r}
df <- ungroup(df) # Ungroup the data to avoid issues with lme
model_lmm <- lme(
  energy ~ batch_size * input_size,
  random = ~ 1 | subject_id / dataset,
  data = df,
  na.action = na.omit
)
summary(model_lmm)

anova(model_lmm)
```

## Post hoc tests
Estimate the effect sizes of the fixed effects in the model.
```{r}
eta_squared(model_lmm, partial = TRUE)
```

```{r}
r2(model_lmm)
```


Estimate marginal means for the interaction between batch size and input size.
```{r}
emm <- emmeans(model_lmm, ~ batch_size * input_size)
summary(emm)
```

Pairwise comparisons for the interaction between batch size and input size.
```{r}
pairs(emm, by = "input_size", adjust = "tukey")
pairs(emm, by = "batch_size", adjust = "tukey")
```

Plot the estimated marginal means.
```{r}
emm_df <- as.data.frame(emm)

ggplot(emm_df, aes(x = batch_size, y = emmean, group = input_size, color = input_size)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(
    title = "Estimated Marginal Means of Energy (Desktop)",
    x = "Batch Size",
    y = "Predicted Energy Consumption",
    color = "Input Size"
  ) +
  theme_minimal()
```
