---
title: "RQ4: How can we accurately estimate energy consumption before training DL models for image classification?"
output: html_notebook
author: "Santiago del Rey"
---

```{r setup, include=FALSE}
library(arrow)
library(lme4)
library(lmerTest) # p-values for lmer
library(DHARMa) # residual diagnostics for mixed models
library(performance)
library(emmeans) # post hoc tests
library(effects)
library(broom)
library(car)
library(patchwork) # for combining plots
library(ggplot2)
library(ggpubr) # for publication-ready plots
library(viridis)
library(showtext)
library(dplyr)

# For discrete variables
viridis_scheme <- "turbo"
options(
  ggplot2.discrete.fill = function(...) scale_fill_viridis_d(..., option = viridis_scheme),
  ggplot2.discrete.colour = function(...) scale_colour_viridis_d(..., option = viridis_scheme)
)

# For continuous variables
options(
  ggplot2.continuous.fill = function(...) scale_fill_viridis_c(..., option = viridis_scheme),
  ggplot2.continuous.colour = function(...) scale_colour_viridis_c(..., option = viridis_scheme)
)

font_add("CMU Serif", regular = "cmunrm.otf")  # You need the font installed
showtext_auto()

theme_set(
  theme_pubr(base_family = "CMU Serif") +
    theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 9),
      legend.text = element_text(size = 9),
      strip.text = element_text(size = 10)
    )
)

fig_dir <- "../reports/figures/RQ4/"
if (!dir.exists(fig_dir)) {
  dir.create(fig_dir, recursive = TRUE)
}
```

## Load data
```{r load-data}
df <- read_parquet("../data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
  rename(
    energy = `energy (MJ)`,
    training_duration = `training duration (h)`,
    used_ram = `average ram used (MB)`,
    batch_size = `batch size`,
    image_size = `image size`,
    allocated_ram = `total ram (GB)`,
    training_environment = `training environment`,
  ) %>%
  filter(architecture == "inception_v3") %>%
  mutate(
    energy_per_image = (energy * 1000000) / `total seen images`, # Convert MJ to J and normalize by total seen images
    training_duration = training_duration * 60, # Convert hours to minutes
    batch_size = factor(batch_size, levels = c("32", "64", "256")),
    input_size = factor(image_size, levels = c("(75, 75)", "(128, 128)", "(224, 224)")),
    allocated_ram = factor(allocated_ram, levels = c("8", "16", "24", "32")),
    dataset = factor(dataset),
  ) %>%
  select(energy_per_image, training_duration, used_ram, batch_size, input_size, dataset, training_environment, allocated_ram)
```

## Visualize the data
### Data summary
```{r data-summary}
df_batch_input <- df %>%
  filter(training_environment == "Server" & allocated_ram == 16) %>%
  group_by(batch_size, input_size, dataset) %>%
  mutate(
    Q1 = quantile(energy_per_image, 0.25),
    Q3 = quantile(energy_per_image, 0.75),
    IQR = Q3 - Q1,
    is_outlier = energy_per_image < (Q1 - 1.5 * IQR) | energy_per_image > (Q3 + 1.5 * IQR)
  ) %>%
  ungroup() %>%
  filter(!is_outlier) %>%
  select(-Q1, -Q3, -IQR, -is_outlier)

df_batch_input %>%
  group_by(batch_size, input_size, dataset) %>%
  get_summary_stats(energy_per_image, type = "mean_sd")

df_allocated_ram_energy <- df %>%
  filter(batch_size == "256" & input_size == "(128, 128)" & dataset == "stanford_dogs" & training_environment == "Server") %>%
  group_by(allocated_ram) %>%
  mutate(
    Q1 = quantile(energy_per_image, 0.25),
    Q3 = quantile(energy_per_image, 0.75),
    IQR = Q3 - Q1,
    is_outlier = energy_per_image < (Q1 - 1.5 * IQR) | energy_per_image > (Q3 + 1.5 * IQR)
  ) %>%
  ungroup() %>%
  filter(!is_outlier) %>%
  select(-Q1, -Q3, -IQR, -is_outlier)

df_allocated_ram_energy %>%
  group_by(allocated_ram) %>%
  get_summary_stats(energy_per_image, type = "mean_sd")

df_allocated_ram_time <- df %>%
  filter(batch_size == "256" & input_size == "(128, 128)" & dataset == "stanford_dogs" & training_environment == "Server") %>%
  group_by(allocated_ram) %>%
  mutate(
    Q1 = quantile(training_duration, 0.25),
    Q3 = quantile(training_duration, 0.75),
    IQR = Q3 - Q1,
    is_outlier = training_duration < (Q1 - 1.5 * IQR) | training_duration > (Q3 + 1.5 * IQR)
  ) %>%
  ungroup() %>%
  filter(!is_outlier) %>%
  select(-Q1, -Q3, -IQR, -is_outlier)
df_allocated_ram_time %>%
  group_by(allocated_ram) %>%
  get_summary_stats(training_duration, type = "mean_sd")
df_allocated_ram_ram <- df %>%
  filter(batch_size == "256" & input_size == "(128, 128)" & dataset == "stanford_dogs" & training_environment == "Server") %>%
  group_by(allocated_ram) %>%
  mutate(
    Q1 = quantile(used_ram, 0.25),
    Q3 = quantile(used_ram, 0.75),
    IQR = Q3 - Q1,
    is_outlier = used_ram < (Q1 - 1.5 * IQR) | used_ram > (Q3 + 1.5 * IQR)
  ) %>%
  ungroup() %>%
  filter(!is_outlier) %>%
  select(-Q1, -Q3, -IQR, -is_outlier)
df_allocated_ram_ram %>%
  group_by(allocated_ram) %>%
  get_summary_stats(used_ram, type = "mean_sd")
```

### Boxplot of energy consumption
```{r energy-boxplot}
ggplot(df_batch_input, aes(x = batch_size, y = energy_per_image, color = input_size)) +
  geom_boxplot() +
  facet_wrap(~dataset) +
  labs(
    title = "Energy Consumption by Batch Size and Input Size",
    x = "Batch Size",
    y = "Energy Consumption (J/image)",
    color = "Input Size"
  )
```

```{r boxplot-allocated-ram, fig.width=15}
allocated_ram_energy <- ggplot(df_allocated_ram_energy, aes(x = allocated_ram, y = energy_per_image)) +
  geom_boxplot() +
  labs(
    x = "Allocated RAM (GB)",
    y = "Energy Consumption (J/image)"
  )

allocated_ram_time <- ggplot(df_allocated_ram_time, aes(x = allocated_ram, y = training_duration)) +
  geom_boxplot() +
  labs(
    x = "Allocated RAM (GB)",
    y = "Training Duration (minutes)"
  )

allocated_ram_ram <- ggplot(df_allocated_ram_ram, aes(x = allocated_ram, y = used_ram)) +
  geom_boxplot() +
  labs(
    x = "Allocated RAM (GB)",
    y = "Average RAM Used (MB)"
  )

ram_impact_plot <- (allocated_ram_energy | allocated_ram_time | allocated_ram_ram) +
  plot_layout(axes = "collect") +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")",
  ) &
  theme(
    plot.tag.position = "bottom"
  )
  # theme(
  #   axis.title = element_text(size = 12),
  #   axis.text = element_text(size = 11),
  # )
ram_impact_plot

ggsave(
  filename = file.path(fig_dir, "ram-impact.pdf"),
  plot = ram_impact_plot,
  device = cairo_pdf,
  units = "cm",
  width = 13.71, height = 9, dpi = 300
)
```

## Statistical analysis of the effects of batch size, input size, and dataset on energy consumption
### Linear model with dataset as a fixed effect
```{r linear-model}
model_lm <- lm(energy_per_image ~ batch_size * input_size * dataset, data = df_batch_input)
summary(model_lm)
```

```{r anova-linear-model}
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_lm) # Type III ANOVA for linear models
```

#### Check assumptions of the linear model
```{r check-dharma}
sim_res <- simulateResiduals(model_lm)
testResiduals(sim_res)
```

### Linear mixed model
```{r linear-mixed-model}
model_lmm <- lmer(
  energy_per_image ~ batch_size * input_size + (1 | dataset),
  data = df_batch_input,
)
summary(model_lmm)
```

```{r anova-mixed-model}
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_lmm, type = 3) # Type III ANOVA for mixed models
```

#### Check assumptions of the linear mixed model
```{r check-dharma-mixed}
sim_res_lmm <- simulateResiduals(model_lmm)
testResiduals(sim_res_lmm)
```

### Generalized linear model with Gamma distribution
```{r generalized-linear-model}
model_glm <- glm(
  energy_per_image ~ batch_size * input_size * dataset,
  data = df_batch_input,
  family = Gamma(link = "log")
)
summary(model_glm)
```

```{r anova-glm}
options(contrasts = c("contr.sum", "contr.poly")) # sum-to-zero contrasts for Type III
Anova(model_glm) # Type III ANOVA for generalized linear models
```

#### Check assumptions of the generalized linear model
```{r check-dharma-glm}
sim_res_glm <- simulateResiduals(model_glm)
testResiduals(sim_res_glm)
```

### Compare models
```{r model-comparison-glm}
compare_performance(model_lm, model_lmm, model_glm, rank = TRUE)
```

### Post hoc tests
Estimate the effect sizes of the fixed effects in the model.
```{r effect-sizes}
# Tidy the model for plotting
tidy_lm <- broom::tidy(model_lm)

# Filter out intercept and sort by absolute effect size
tidy_lm_filtered <- tidy_lm[tidy_lm$term != "(Intercept)", ]
tidy_lm_filtered <- tidy_lm_filtered[order(abs(tidy_lm_filtered$estimate), decreasing = TRUE), ]
# Filter out terms with NA estimates
tidy_lm_filtered <- tidy_lm_filtered %>%
  filter(!is.na(estimate))

# Plot
eff_plot <- ggplot(tidy_lm_filtered, aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Effect Sizes",
    x = "Term",
    y = "Estimate (Energy Consumption Impact)"
  ) +
  theme_pubr()
```

Estimate marginal means for the interaction between batch size and input size, conditioned on the dataset.
```{r post-hoc}
emm_lm <- emmeans(model_lm, ~ batch_size * input_size | dataset)
emm_lm
```

```{r post-hoc-pairs}
pairs(emm_lm, by = c("dataset", "batch_size"), adjust = "tukey")
```

Plot the estimated marginal means.
```{r fig.height=8, fig.width=15}
emm_df <- as.data.frame(emm_lm) %>%
  filter(!is.na(emmean))

emm_plot <- ggplot(emm_df, aes(x = batch_size, y = emmean, group = input_size, color = input_size)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~dataset) +
  labs(
    title = "Estimated Marginal Means of Energy",
    x = "Batch Size",
    y = "Predicted Energy Consumption",
    color = "Input Size"
  ) +
  theme_pubr()

combined_plot <- (eff_plot | emm_plot) +
  plot_annotation(title = "Energy Consumption Analysis", theme = theme(plot.title = element_text(hjust = 0.5)))

combined_plot
```

## Statistical analysis of the effects of allocated RAM on energy consumption
### Linear model with allocated RAM as a fixed effect
```{r linear-model-allocated-ram}
model_lm_ram <- lm(energy_per_image ~ allocated_ram, data = df_allocated_ram_energy)
summary(model_lm_ram)
```
```{r anova-linear-model-allocated-ram}
anova(model_lm_ram) # Type I ANOVA for linear models
```
#### Check assumptions of the linear model with allocated RAM
```{r check-dharma-allocated-ram}
sim_res_ram <- simulateResiduals(model_lm_ram)
testResiduals(sim_res_ram)
```

#### Post hoc tests for allocated RAM
```{r post-hoc-allocated-ram}
emm_ram <- emmeans(model_lm_ram, ~allocated_ram)
emm_ram
pairs(emm_ram, adjust = "tukey")

eff_size(emm_ram, sigma = sigma(model_lm_ram), edf = df.residual(model_lm_ram))

plot(emm_ram, comparisons = TRUE) +
  labs(
    title = "Estimated Marginal Means of Energy by Allocated RAM",
    x = "Predicted Energy Consumption (J/image)",
    y = "Allocated RAM (GB)",
  ) +
  theme_pubr()
```

## Statistical analysis of the effects of allocated RAM on training duration
### Linear model with allocated RAM as a fixed effect
```{r linear-model-training-duration}
model_lm_duration <- lm(training_duration ~ allocated_ram, data = df_allocated_ram_time)
summary(model_lm_duration)
anova(model_lm_duration) # Type I ANOVA for linear models
```
#### Check assumptions of the linear model with allocated RAM
```{r check-dharma-training-duration}
sim_res_duration <- simulateResiduals(model_lm_duration)
testResiduals(sim_res_duration)
```
#### Post hoc tests for allocated RAM on training duration
```{r post-hoc-training-duration}
emm_duration <- emmeans(model_lm_duration, ~allocated_ram)
emm_duration
pairs(emm_duration, adjust = "tukey")
eff_size(emm_duration, sigma = sigma(model_lm_duration), edf = df.residual(model_lm_duration))
plot(emm_duration, comparisons = TRUE) +
  labs(
    title = "Estimated Marginal Means of Training Duration by Allocated RAM",
    x = "Predicted Training Duration (minutes)",
    y = "Allocated RAM (GB)",
  ) +
  theme_pubr()
```

## Statistical analysis of the effects of allocated RAM on average RAM used
### Linear model with allocated RAM as a fixed effect
```{r linear-model-average-ram-used}
model_lm_ram_used <- lm(used_ram ~ allocated_ram, data = df_allocated_ram_ram)
summary(model_lm_ram_used)
anova(model_lm_ram_used) # Type I ANOVA for linear models
```

#### Check assumptions of the linear model with allocated RAM
```{r check-dharma-average-ram-used}
sim_res_ram_used <- simulateResiduals(model_lm_ram_used)
testResiduals(sim_res_ram_used)
```
#### Post hoc tests for allocated RAM on average RAM used
```{r post-hoc-average-ram-used}
emm_ram_used <- emmeans(model_lm_ram_used, ~allocated_ram)
emm_ram_used
pairs(emm_ram_used, adjust = "tukey")
eff_size(emm_ram_used, sigma = sigma(model_lm_ram_used), edf = df.residual(model_lm_ram_used))
plot(emm_ram_used, comparisons = TRUE) +
  labs(
    title = "Estimated Marginal Means of Average RAM Used by Allocated RAM",
    x = "Predicted Average RAM Used (MB)",
    y = "Allocated RAM (GB)",
  ) +
  theme_pubr()
```
