---
title: "RQ4: How can we accurately estimate energy consumption before training DL models for image classification?"
output: html_notebook
author: "Santiago del Rey"
---

## Load libraries
```{r setup, include=FALSE}
library(arrow)
library(dplyr)
library(lme4)
library(nlme)
library(lmerTest) # p-values for lmer
library(MuMIn)
library(DHARMa) # residual diagnostics for mixed models
library(performance) # check assumptions
library(emmeans) # post hoc tests
library(effects)
library(broom)
library(car)
library(rstatix)
library(effectsize)
library(patchwork) # for combining plots
library(ggplot2)
library(ggpubr) # for publication-ready plots
library(tikzDevice)
```

## Load data
```{r load-data}
df <- read_parquet("../data/metrics/processed/clean-dl-training-energy-consumption-dataset.gzip") %>%
  rename(
    training_environment = `training environment`,
    energy = `energy (MJ)`,
    batch_size = `batch size`,
    image_size = `image size`,
    allocated_ram = `total ram (GB)`,
  ) %>%
  filter(architecture == "inception_v3") %>%
  mutate(
    energy_per_image = (energy * 1000000) / `total seen images`, # Convert MJ to J and normalize by total seen images
    input_size = factor(image_size, levels = c("(75, 75)", "(128, 128)", "(224, 224)")),
    dataset = factor(`dataset`),
  ) %>%
  select(energy_per_image, batch_size, input_size, dataset, training_environment, allocated_ram)
```

## Visualize the data
### Data summary
```{r data-summary}
df_batch_input <- df %>%
  filter(training_environment == "Server" & allocated_ram == 16) %>%
  group_by(batch_size, input_size) %>%
  mutate(
    Q1 = quantile(energy_per_image, 0.25),
    Q3 = quantile(energy_per_image, 0.75),
    IQR = Q3 - Q1,
    is_outlier = energy_per_image < (Q1 - 1.5 * IQR) | energy_per_image > (Q3 + 1.5 * IQR)
  ) %>%
  ungroup() %>%
  filter(!is_outlier) %>%
  select(-Q1, -Q3, -IQR, -is_outlier)

df_batch_input %>%
  group_by(batch_size, input_size, dataset) %>%
  get_summary_stats(energy_per_image, type = "mean_sd")

df_allocated_ram <- df %>%
  filter(batch_size == "256" & input_size == "(128, 128)" & dataset == "stanford_dogs" & training_environment == "Server") %>%
  group_by(allocated_ram) %>%
  mutate(
    Q1 = quantile(energy_per_image, 0.25),
    Q3 = quantile(energy_per_image, 0.75),
    IQR = Q3 - Q1,
    is_outlier = energy_per_image < (Q1 - 1.5 * IQR) | energy_per_image > (Q3 + 1.5 * IQR)
  ) %>%
  ungroup() %>%
  filter(!is_outlier) %>%
  select(-Q1, -Q3, -IQR, -is_outlier)

df_allocated_ram %>%
  group_by(allocated_ram) %>%
  get_summary_stats(energy_per_image, type = "mean_sd")
```

### Boxplot of energy consumption
```{r energy-boxplot}
ggboxplot(
  df_batch_input,
  x = "batch_size", y = "energy_per_image",
  color = "input_size", palette = "jco"
) +
  facet_wrap(~dataset) +
  labs(
    title = "Energy Consumption by Batch Size and Input Size",
    x = "Batch Size",
    y = "Energy Consumption (J/image)",
    color = "Input Size"
  ) +
  theme_pubr()

ggboxplot(
  df_allocated_ram,
  x = "allocated_ram", y = "energy_per_image", palette = "jco"
)
```

## Statistical analysis of the effects of batch size, input size, and dataset on energy consumption
### Linear model with dataset as a fixed effect
```{r linear-model}
model_lm <- lm(energy_per_image ~ batch_size * input_size * dataset, data = df_batch_input)
summary(model_lm)
anova(model_lm)
```

#### Check assumptions of the linear model
```{r check-dharma}
sim_res <- simulateResiduals(model_lm)
testResiduals(sim_res)
```

```{r check-assumptions, fig.width=10, fig.height=10}
check_model(model_lm)
```


### Linear mixed model
```{r linear-mixed-model}
model_lmm <- lmer(
  energy_per_image ~ batch_size * input_size + (1 | dataset),
  data = df_batch_input,
)
summary(model_lmm)

anova(model_lmm)
```

#### Check assumptions of the linear mixed model
```{r check-dharma-mixed}
sim_res_lmm <- simulateResiduals(model_lmm)
testResiduals(sim_res_lmm)
```
```{r check-assumptions-mixed, fig.width=10, fig.height=10}
check_model(model_lmm)
```

### Generalized linear model with Gamma distribution
```{r generalized-linear-model}
model_glm <- glm(
  energy_per_image ~ batch_size * input_size * dataset,
  data = df_batch_input,
  family = Gamma(link = "log")
)
summary(model_glm)
anova(model_glm)
```

#### Check assumptions of the generalized linear model
```{r check-dharma-glm}
sim_res_glm <- simulateResiduals(model_glm)
testResiduals(sim_res_glm)
```
```{r check-assumptions-glm, fig.width=10, fig.height=10}
check_model(model_glm)
```

### Compare models
```{r model-comparison-glm}
compare_performance(model_lm, model_lmm, model_glm, rank = TRUE)
```

### Post hoc tests
Estimate the effect sizes of the fixed effects in the model.
```{r effect-sizes}
# Tidy the model for plotting
tidy_lm <- broom::tidy(model_glm)

# Filter out intercept and sort by absolute effect size
tidy_lm_filtered <- tidy_lm[tidy_lm$term != "(Intercept)", ]
tidy_lm_filtered <- tidy_lm_filtered[order(abs(tidy_lm_filtered$estimate), decreasing = TRUE), ]
# Filter out terms with NA estimates
tidy_lm_filtered <- tidy_lm_filtered %>%
  filter(!is.na(estimate))

# Plot
eff_plot <- ggplot(tidy_lm_filtered, aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Effect Sizes",
    x = "Term",
    y = "Estimate (Energy Consumption Impact)"
  ) +
  theme_pubr()
```

Estimate marginal means for the interaction between batch size and input size, conditioned on the dataset.
```{r post-hoc}
emm_lm <- emmeans(model_glm, ~ batch_size * input_size | dataset)
emm_lm
```

```{r post-hoc-pairs}
pairs(emm_lm, by = c("dataset", "input_size"), adjust = "tukey")
pairs(emm_lm, by = c("dataset", "input_size"), adjust = "tukey")
```

Plot the estimated marginal means.
```{r fig.height=8, fig.width=15}
emm_df <- as.data.frame(emm_lm) %>%
  filter(!is.na(emmean))

emm_plot <- ggplot(emm_df, aes(x = batch_size, y = emmean, group = input_size, color = input_size)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~dataset) +
  labs(
    title = "Estimated Marginal Means of Energy",
    x = "Batch Size",
    y = "Predicted Energy Consumption",
    color = "Input Size"
  ) +
  theme_pubr()

combined_plot <- (eff_plot | emm_plot) +
  plot_annotation(title = "Energy Consumption Analysis", theme = theme(plot.title = element_text(hjust = 0.5)))

combined_plot
```
